### 2025-02-25 manuscript/main.tex
Added manuscript latex file

%! BibTeX Compiler = biber
\documentclass{article}
\usepackage{caption}
\usepackage{censor}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
linkcolor=DARKBLUELINK,
anchorcolor=DARKBLUELINK,
citecolor=DARKBLUELINK,
filecolor=DARKBLUELINK,
menucolor=DARKBLUELINK,
urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[
    style=authoryear,
    doi=false,
    isbn=false,
    url=false,
    date=year,
    natbib=true,
    hyperref,
    mincitenames = 1,
    maxcitenames = 2,
    minbibnames = 1,
    maxbibnames = 30,
    uniquename=false,
    uniquelist=false,
    giveninits=true,
    backend=biber]{biblatex}
\AtEveryBibitem{\clearfield{note}}
\addbibresource{references_bibtex.bib}

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}
\usepackage{tabu}
\usepackage{hhline}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{marvosym}
\usepackage{orcidlink}

% -- Defining colors:
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
\lstdefinestyle{mystyle}{
backgroundcolor=\color{backcolour},
basicstyle=\ttfamily\scriptsize\bfseries,
breakatwhitespace=false,
breaklines=true,
captionpos=t,
keepspaces=true,
showspaces=false,
showstringspaces=false,
showtabs=false,
tabsize=2
}% -- Setting up the custom style:
\lstset{style=mystyle}
\captionsetup[table]{hypcap=false}
\captionsetup[figure]{hypcap=false}

\newcommand{\Multiply}{\cdot}
\DeclareMathOperator{\Var}{\text{var}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\PhyloSupport}{\pi}
\newcommand{\MeanPhyloSupport}{\overline{\PhyloSupport}}
\newcommand{\Indiv}{k}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\MutationRatePheno}{\mu}
\newcommand{\MutationRateNuc}{u}
\newcommand{\NbrLoci}{L}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\brownian}{\mathcal{B}}

\renewcommand{\baselinestretch}{1.5}
\renewcommand{\arraystretch}{1.2}
\linenumbers
\frenchspacing

% \title{Phylograms disentangle drift and selection acting on a continuous trait.}
% \title{Drift and selection acting on a continuous trait can be disentangled by a simultaneous use of a chronogram and a phylogram}
% \title{Phylogram instead of chronogram when claiming that a trait is evolving under drift.}
\title{Phylogram for testing neutral evolution of a trait, chronogram to model selection.}
\rhead{\scshape Chronogram or phylogram for trait evolution.}

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\begin{document}
%TC:ignore
\maketitle

% Abstract (≤ 300 words)
\begin{abstract}
    The evolution of traits across different species is driven by a combination of selective and neutral processes, while their relative importance is subject to numerous debates.
    To disentangle them, different scenarios of evolution are modelled and compared.
    Typically, a neutrally evolving trait is modelled as changing randomly along the different branches of the species tree.
    This species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
    Theoretically, changes for a neutral trait depend on the number of generations passed, which in turn correlates with time, hence the use of chronogram.
    But, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases.
    Instead, for a phylogram the branch lengths represent sequence divergence (i.e. the number of substitutions), which absorb the effect of changing generation time and mutation rates.
    In this study, we first show using simulations that for a trait evolving neutrally the fit of a Brownian process has more support on a phylogram than on a chronogram.
    However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions.
    Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift.
    Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a trait.
\end{abstract}

\keywords{Trait evolution \and Phylogenetics \and Phylogenetic Comparative Methods \and Chronogram \and Sequence divergence \and Neutral evolution}

%TC:endignore
\section*{Lay summary}\label{sec:summary}
When we look at different species, we can see that some traits are different between them.
For example, the size of the brain or the expression level of a gene varies between species.
The question we ask is: do the time that have passed since the species diverged explain the difference in traits?
Or is it the genetic differences between species that explains best the differences?
The standard in evolutionary biology is to use the time to explain the differences in traits.
However, if the trait is evolving neutrally, the genetic differences between species instead of time should explain best the differences in traits.
As a result, we argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.

\section*{Introduction}\label{sec:introduction}
% The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
Across species, distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed variation.
Such regimes of evolution are typically compared using phylogenetic comparative methods, where a continuous traits is modeled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
For example, to model neutral evolution, the mean trait value is said to follow a Brownian Motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
Or in other words, for each  branch of the tree the value at descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
In this framework, reconstructing the trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.

% Alternative models of trait evolution and their pitfalls.
Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, when trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
Methodologically, such alternative models of evolution raise issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{uyeda_novel_2014, grabowski_cautionary_2023}.
Finally, the optimum can also be allowed to change continuously along the tree, which is interpreted as a trait tracking a moving optimum~\citep{hansen_translating_1996, latrille_detecting_2024}.
One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
This modeling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a fluctuating optimum.

% This comes from the fact that we use chronogram, and not phylogram.
% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}).
Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
Theoretically, for a neutrally evolving trait, trait changes depends directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the generation time is constant.
But, since species generation time might vary along the phylogenetic tree, modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
Instead, on a phylogram, the branch lengths can represent another quantity than time.
For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction~\citep{wilson_chronogram_2022}.
From a genomic perspective, phylogram can also be obtained in unit of nucleotide divergence, which is equivalent to measuring the number of nucleotide substitutions occurring along a branch.
Such a phylogram in unit of nucleotide divergence obtained from neutrally evolving genomic loci would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
Conversely, for a trait evolving under selection, using a chronogram would be more accurate since changes in the optimum fitness peak is extrinsic to the species and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
% Finally, from a statistical perspective, using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks) to estimate the branch lengths of chronograms.

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_rational}
    \caption{
        Panel~A: Across many species, the evolution of continuous traits can be modeled as a stochastic process evolving along the branches of a phylogenetic tree.
        The branches of such phylogenetic tree can be measured in either time (chronogram) or number of substitutions (phylogram).
        Panel~B: Theoretically, changes of a trait under a moving optimum should be better predicted by a chronogram.
        Instead, changes of a neutral trait should be better predicted a phylogram.
    }
    \label{fig:methods}
\end{figure*}
%TC:endignore

% Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
In this study, we test if it is more accurate to use a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram as the underlying tree to model neutral trait changes, by evaluating by the fit of a BM on both trees (Fig.~\ref{fig:methods}B).
In other words, we seek to test if change of a neutral trait is better predicted a phylogram.
Conversely, we also seek to test if changes of a trait under a moving optimum is better predicted by a chronogram.
Altogether, by using genomic information we seek to provide additional approaches to model trait evolution.

\section*{Methods}
\subsection*{Simulations along a phylogenetic tree}
We performed simulations under different selective regimes (neutral, multiple optima, moving optimum).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
The parameters of simulations were chosen to mimic an empirical dataset of \textit{Mammal}s.
In summry, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
Parents were selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
Flattening the fitness landscape result in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch, if a switch occurs the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).

At each node of the tree, the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached (Fig.~\ref{fig:results-brownian}C).
As a control, we performed simulations with constant mutation rate, generation times and effective population sizes ($\Ne$).
Alternatively, we performed simulations with fluctuating mutation rate, generation times and $\Ne$, were we used a Brownian process to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).

The phylogram is obtained from the same simulation setting on a set of independent 30.000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood smoothing parameter of 1) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

\subsection*{Empirical dataset}
We analyzed a dataset of body and brain masses from \textit{Mammal}s.
The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
We also extracted brain and body masses from the \textit{COMBINE} dataset~\citep{soria_combine_2021}.
The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
More specifically, the phylogram in unit of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood smoothing parameter of 1) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

\subsection*{Brownian Motion (BM)}
We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian Motion (BM) on continuous traits evolution evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
The data consist of the mean trait value for each extant species, and the tree topology is fixed.
Along each branch, the value of the trait is drawn from a normal distribution with mean equal to the parent node value and variance equal to the branch length~\citep{felsenstein_phylogenies_1985}.
We reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
Both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
Importantly, the data and the tree topology are the same in both analyses, only the branch lengths are different.

Additionally to a single rate BM, we fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1000 gen., running of 50000 gen., 2 chains).

Finally, we implemented a model in \textit{RevBayes} containing a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
As in the previous section, a Brownian process is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length.
Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favored over the chronogram.

\subsection*{Ornstein-Uhlenbeck (OU)}
% Simple OU model versus a brownian process.
First, we compared the fit of an Ornstein-Uhlenbeck (OU) process (eq.~\ref{eq:ou-process}) to a BM~\citep{hansen_stabilizing_1997,butler_phylogenetic_2004}, using either a phylogram or a chronogram as the underlying tree.
% Ornstein-Uhlenbeck model equations
\begin{align}
    \der \Trait_t & = -\alpha \left( \Trait_t - \theta \right) \der t + \sigma \der W_t \label{eq:ou-process}
\end{align}
In the OU process, the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, and $\theta$ is the optimum value.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the Brownian process, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
% alpha ~ dnReversibleJumpMixture(0.0, dnExponential( abs(root_age / 2.0 / ln(2.0)) ), 0.5)
Mathematically, $\alpha$ is drawn from a reversible-jump mixture distribution: a value of $0$ or drawn from an exponential distribution.
Thus, when $\alpha = 0$ the OU process is equivalent to a BM.
The support for the OU model is estimated as the posterior mean of the switch variable, or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).

% Relaxed OU model, how many optimums?
Second, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1000 gen., running of 5000 gen., 2 chains).

\section*{Results}

\subsection*{Support for a phylogram over a chronogram}
% Simulations of a trait evolving neutrally versus under a moving optimum.
We developed a Bayesian model to estimate the support of phylogram over a chronogram for a trait evolving under a Brownian Motion (BM) process (see Methods, BM section).
To test this model, we simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) evolving under multiple optima, and 3) evolving under a moving optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
As a control experiment we simulated with a constant generation time, mutation rate (per generation) and effective population size ($\Ne$).
In such a case, the support is the same for both a phylogram and a chronogram (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
Next, we simulated changing generation time, mutation rate and $\Ne$ along the phylogenetic tree mimicking a mammalian range of changes.
In this case, the BM was better fitted on a phylogram than on a chronogram for a neutral trait (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
Conversely, the fit of a BM on a chronogram was better than on a phylogram for a trait under selection, with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).

%TC:ignore
\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_simulations}\\
    \vspace{10pt}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small  \textbf{D: simulations with constant generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_cstGT_phylogram_support}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small  \textbf{E: simulations with fluctuating generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_phylogram_support}
    \end{minipage}
    \caption{
        Panel~A: Wright-Fisher simulations with mutation, selection and drift.
        For a given individual, its genotypic value is encoded by a finite number of independent loci (meaning no linkage) contributing additively.
        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
        Panel~B: Drift is modeled by the resampling of parents.
        Selection is modeled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree.
        Panel~C: Example simulation of a trait evolving along a phylogenetic tree.
        Panels D \& E: Violinplot of posterior probability of the phylogram being favored over the chronogram for the different simulated regime of selection.
        Horizontal lines inside the violin show the result of each replicate simulation.
        Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow), under a moving optimum (blue), and under multiple optima (red).
        Results of simulations with constant generation time (Panel~D) and with fluctuating generation time, mutation rate and effective population size (Panel~E).
    }
    \label{fig:results-brownian}
\end{figure*}
%TC:endignore

\subsection*{Trait evolution on a phylogram and on a chronogram}
% Mis-classification of trait evolution on a phylogram.
Beside support for the underlying tree, ancestral state reconstruction can also be improved by the use of a chronogram or a phylogram.
For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$)
than a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).
Additionally, instead of testing the fit of a BM with a single rate, we also fitted a multi-rate BM process (see Methods).
When fitting a multi-rate BM process, the estimated variance of rate parameters ($v$) was lower on a phylogram than on a chronogram for a trait evolving neutrally (Fig.~\ref{fig:results-alternative}A, yellow violins, Wilcoxon paired test with $p_{\\text{value}}=8.8\times 10^{-14}$), and the number of rate shifts ($n$) was reflecting the prior on the phylogram while not on the chronogram, showing a bias (Fig.~\ref{fig:results-alternative}B, yellow violins).
Conversely, for a trait under a moving optimum, $v$ was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}A, blue violins, Wilcoxon paired test with  $p_{\\text{value}}=8.9\times 10^{-15}$), and $n$ was reflecting the prior on the chronogram while not on the phylogram, showing a bias (Fig.~\ref{fig:results-alternative}B, blue violins).
Altogether, fitting a BM on a phylogram is more accurate for a trait evolving neutrally, but results in less accurate estimates for a trait under selection.

% Mainly testing BM versus OU
Beside testing the fit of a BM, we tested alternative Ornstein-Uhlenbeck (OU) models: with a single optimum and multi-optima (see Methods).
First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\\text{value}}=3.9\times 10^{-18}$).
Worst still, even for a trait evolving neutrally, $p_{\\text{OU}}$ was still higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, yellow violins, Wilcoxon paired test with $p_{\\text{value}}=8.0\times 10^{-8}$).
When fitting a multi-OU process on a trait evolving neutrally, the number of optimum shifts ($m$) was reflecting the prior on the phylogram while not on the chronogram (Fig.~\ref{fig:results-alternative}B, yellow violins).
Conversely, for a trait under a moving optimum, ($m$) was reflecting the prior on the chronogram while not on the phylogram (Fig.~\ref{fig:results-alternative}A, blue violins).
Altogether, fitting an OU process on a phylogram is more accurate for a trait evolving under selection, but results in less accurate estimates for a trait evolving neutrally.

%TC:ignore
\begin{figure*}[!ht]
    \centering
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{A}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_MultiBM}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{B}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_MultiBM_NSwitch}
    \end{minipage}\\
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{C}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_OU}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{D}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_MultiOU_NSwitch}
    \end{minipage}

    \caption{
        \textbf{Mis-classification of trait evolution on a phylogram.}
        Violinplot of posterior parameter estimates for different models of trait evolution and different simulated regime of selection.
        Horizontale lines inside the violin show the result of each replicate simulation.
        Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow) and under a moving optimum (blue).
        Wilcoxson rank-test are performed between the paired estimates on the phylogram and the chronogram.
        Panel~A \& B: Fit of a relaxed Brownian Motion (BM) with multiple rate parameters, using either a phylogram or a chronogram.
        Posterior estimate for the variance in rate parameters (Panel~A) and number of rate changes (Panel~B).
        Panel~C: Relative fit an Ornstein-Uhlenbeck (OU) process compared to BM, using either a phylogram or a chronogram.
        Posteriors estimates for the support of the OU process over the BM (reversible-jump).
        Panel~D: Fit of a relaxed OU with multiple optima, using either a phylogram or a chronogram.
        Posterior estimate for the number of optimum changes.
    }
    \label{fig:results-alternative}
\end{figure*}
%TC:endignore

\section*{Discussion}
% Context and rationale
In phylogenetic comparative methods, the evolution of continuous traits is typically modeled as a stochastic process running along the branches of a phylogenetic tree.
Such branches are measured in either time (tree is a chronogram) or in number of nucleotide substitutions (tree is a phylogram).
Typically, the null model of neutral trait evolution (i.e.~drift) is though as a Brownian process running on a chronogram~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
As a result, dataset on which the Brownian process is favored over alternative models are often interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
However, we argue instead that claims for a trait evolving under drift should be tested more thoroughly by comparing the fit of the Brownian process under both a phylogram and a chronogram as the underlying tree.
Indeed, for a trait evolving neutrally, changes of generation time along the tree would be absorbed by the phylogram while not by the chronogram~\citep{latrille_detecting_2024}, so we argue that the chronogram should not be used as the backbone tree in this case.
Moreover, using phylogram would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks) to estimate divergence times of chronograms.

% Our main result
% Claiming that a trait is evolving under drift should be tested by comparing the fit under a phylogram and under a chronogram.
We developed a Bayesian model implemented in \textit{RevBayes}~\citep{hohna_revbayes_2016} to indicate whether a Brownian process is fitted better on a phylogram or a chronogram ($\PhyloSupport$), given data at the tip of the tree.
For simulations of a trait evolving neutrally, we showed that the fit of a Brownian process supports a phylogram rather than a chronogram.
Moreover, for the neutral trait, both ancestral state reconstruction and estimation for the number of rate changes for a multi-rate BM was also more accurate on a phylogram.
Conversely, for simulation under a moving optimum, the fit of a Brownian process on a chronogram was better than on a phylogram for a trait.
Applied to different empirical dataset of body and brain masses in mammals, we showed that the chronogram was favored over the phylogram, ruling out the neutral model of evolution for these trait~\citep{latrille_detecting_2024}.

% When does it apply and why is it important
The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree.
But more generally, our method is also relevant when changes in mutation rate (per generation) and effective population size ($\Ne$) are expected.
First, normalization by nucleotide divergence also accounts theoretically for variations in effective population size ($\Ne$) since the substitution rate of neutral mutations is equal to the mutation rate, such that $\Ne$ has no effect on the rate of neutral evolution~\citep{kimura_evolutionary_1968, ohta_population_1972}.
Moreover, if population structure changes of the probability of fixation of neutral mutations (that would also impact a neutral trait), the phylogram would absorb automatically these changes.
This argument is not only verbal since we tested for it in our simulation by modelling changes in both $\Ne$ and generation time, and the phylogram was favored over the chronogram for a trait evolving neutrally.
Second, aside from absorbing of changes in generation time and $\Ne$, the use of phylograms also absorbs changes in mutation rate (per generation) under certain assumptions.
As is the case in our simulations, we assumed that the genetic architecture of the trait is constant and the mutation rate is homogeneous across the genome.
However, if the mutation rate would only changes for a subset of the genome that is encoding the trait, the phylogram would not absorb these changes in mutation rate.

% When does it fall appart, our second result
These examples show that while phylograms are useful to neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
In our simulations, we found that generally using a phylogram tend to favor the OU process over the Brownian Motion (BM) model.
This suggests that trees in unit of nucleotide divergence may not always provide a reliable framework for modeling trait evolution, especially when selection is involved.
As such, we recommend continuing to use chronograms for modeling selection acting on a trait, as they generally provide more accurate and reliable results.
Moreover, for a trait is under selection for which the pace of optimum changes is also tracking the changes in generation time, the support for a phylogram over a chronogram would be misleading~\citep{litsios_effects_2012}.
Therefore, we argue that while phylograms can offer valuable insights, they should be used with caution and in conjunction with chronograms to provide a more comprehensive understanding of trait evolution.

% Altogether, what we have done and haven't tested and done yet
Our results suggest that the use of phylograms can be a valuable tool for modeling trait evolution, particularly when the null model is neutral evolution.
First, we simulated only a subset of different regimes of selection, but more complex scenarios could be included for example with with primary and secondary moving optimums~\citep{hansen_three_2024}.
Additionally, the simulated genetic architecture of a single trait is assumed to be constant across the phylogeny.
Such constant architecture result in saturation of the phenotype on long time scales, with a phenomenon similar to nucleotide saturation~\citep{latrille_detecting_2024}.
Such saturation might be one of the reason for a bias in favor of the Ornstein-Uhlenbeck (OU) process over the Brownian Motion (BM) model that we observed in our simulations.
A more realistic genetic architecture should include pleiotropy, epistasis and linkage disequilibrium, while here we only considered non-linked loci contributing additively to a single trait.
Second, while the tested alternative models of evolution (multi-rates BM, OU, multi-optium OU) are standard, they do not encompass the full breadth of available methods, which could be expanded~\citep{pennell_geiger_2014, hohna_revbayes_2016}.
Finally, the dataset of body and brain masses in \textit{Mammal}s may not be representative of all traits or species and even though our analysis showcase the utility of phylograms, more empirical studies are needed to replicate the findings.

% Gene expression level and empirical studies.
One striking example of traits that could (and should) be studied with a phylogram is gene expression level, both for protein concentration and mRNA expression.
Indeed, the regime of selection acting on gene expression level is the focus of intense debates~\citep{signor_evolution_2018, price_detecting_2022, bertram_cagee_2023, dimayacyac_evaluating_2023}.
And typically, dataset on which the Brownian process is favored over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
More generally, empirical studies plot phenotypic distance as a function of time while claiming that a trait is evolving under drift~\citep{jiang_decoupling_2023}.
As such, we argue that phenotypic distance should be computed as the function of branch length in unit of substitutions (square root), not time.
Altogether, the use of a phylogram, or more generally nucleotide divergence could help to disentangle the effect of drift and selection acting on gene expression levels.

% Perspective on the use of phylograms in phylogenetic comparative methods.
The use of phylograms in phylogenetic comparative methods has been a subject of debate~\citep{wilson_chronogram_2022}.
The study of \citet{wilson_chronogram_2022} suggested that phylograms are more robust for studying discrete character evolution, contrary to our results suggesting more false positives.
However, this apparent contradiction is due to the different units of measurement used in the studies, as they measure branch lengths in units of morphological distance rather than substitutions per site.
We argue that the use of a phylogram, measured in units of substitutions, could also be useful for studying the evolution of discrete characters~\citep{wilson_chronogram_2022}.
Phylograms provide a valuable tool for modeling trait evolution, especially when the null model is neutral evolution.
In addition to the neutrality index for a trait~\citep{latrille_detecting_2024}, the toolbox for studying trait evolution could be enriched by the use of phylograms to model trait evolution, particularly when variation within species is available.
We argue, that the soundness of studying trait evolution on phylograms can be evaluated by the absolute fit of a model to the data~\citep{pennell_model_2015}.
More generally, genomic information could potentially be seen as a way to disentangle congruence models~\citep{louca_extant_2020}, or as prior for methods that detect shifts in adaptive regimes~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020}.

% Conclusion
Altogether, we support the use a chronogram when testing for selection acting on a trait.
The pipeline of hypothesis testing and model comparison is currently well established and should be followed.
However, when the analysis concludes that the BM is favored, then the support of the phylogram over the chronogram should be tested additionally.
We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis.
Only then we can we have a bit more support that a trait is evolving under drift if the phylogram has more support than chronogram, but caution is still warranted.

%TC:ignore
\printbibliography

\section*{Author contributions}
Original idea: T.L.\ and anonymous reviewers;
Model conception: T.L., T.G.\ and N.S.;
Code: T.L.;
Data analyses: T.L.\ and T.G.;
Interpretation: T.L., T.G.\ and N.S.;
First draft: T.L.;
Editing and revisions: T.L., T.G.\ and N.S.
Project management and funding: N.S\@.

\section*{Acknowledgements}
\label{sec:acknowledgment}
We gratefully acknowledge the help of Julien Joseph for his advice and reviews concerning this manuscript.

\section*{Funding}
Université de Lausanne

\section*{Competing interests}
The authors declare no conflicts of interest.

\subsection*{Data and materials availability:}
The materials that support the findings of this study are openly available in GitHub at \href{https://github.com/ThibaultLatrille/ChronoPhylogram}{github.com/ThibaultLatrille/ChronoPhylogram}.
Snakemake pipeline, simulator, analysis scripts and \textit{RevBayes} scripts and documentation are available in the repository to replicate the study.

\newpage

\part*{Supplementary materials}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{section}{0}

\renewcommand{\baselinestretch}{1.0}\normalsize
\tableofcontents
\renewcommand{\baselinestretch}{1.5}\normalsize

\section{Simulation of continuous trait}\label{sec:simulator}

\subsection{Genotype-phenotype map}\label{subsec:genotype-phenotype-map}

\begin{itemize}
    \item $\NbrLoci$ is the number of loci encoding the trait.
    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
    \item $\Ne$ is the effective number of individuals.
    \item $g_{\Indiv,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $\Indiv \in \{1, \hdots, \Ne\}$.
    \item $G_{\Indiv} = \sum_{l=1}^{\NbrLoci} a_l \times g_{\Indiv,l}$ is the genotypic value for individual $\Indiv$.
    \item $\xi_{\Indiv} \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual $\Indiv$.
    \item $\Trait_{\Indiv} = G_{\Indiv}+ \xi_i$ is the phenotype for individual $\Indiv$.
\end{itemize}

\begin{center}
    \captionof{figure}{summary of trait's genetic architecture.}
    \includegraphics[width=0.6\textwidth, page=1] {figures/artwork_genet_architecture}
    \label{fig:simulator-summary}
\end{center}

\subsection{Simulation along a tree}\label{subsec:simulation-along-a-tree}
% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
% pS = 0.0027577478571428568
% Tree length (dS) = 2.340579
% Tree length (My) = 1259.4490200000002
% Root age (My) = 98.9489413888889
% Root age (dS) = 0.18388820079995308
% nbr_sites_var = 27.577478571428568
% u = 1.3788739285714283e-05
% nbr_generations = 13336.114128321291
Each individual phenotypic value was the sum of genotypic value and an environmental effect.
The environmental effect was normally distributed with variance $\VarEnv$.
We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$.
We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly.
Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRatePheno \Multiply \NbrLoci $, with the mutation rate per locus per generation $\MutationRatePheno$.
From the empirical mammalian dataset (see Methods), we computed an average nucleotide divergence from the root to leaves of $0.18$.
We scaled parameters in our simulations to fit plausible values for \textit{Mammal}s.
We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per site per generation and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
We set $\MutationRatePheno=\MutationRateNuc$ without loss in generality since the genetic architecture ($\NbrLoci$ and $a$) is assumed constant in the simulator.

The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modeled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0043\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0043\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0043 \Multiply \sqrt {13,500} = 0.5$.
In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $0.5$.
An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
From a speciation process perspective, this was equivalent to an allopatric speciation over one generation.

At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
Mutations were considered as a displacement of the phenotype in the multidimensional space.
Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
Moving optimum selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
Multiple optima were implemented by allowing the optimum phenotype to shift along a branch with a probability of $10^{-5}$, the shift in optimum being then drawn from a exponential distribution with rate parameter $0.1$, meaning an average of $10$ in the optimum shift per switch.
Neutral evolution was implemented by flattening the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.

\newpage

\section{Empirical dataset}\label{sec:empirical-dataset}

\begin{center}
\begin{adjustbox}{width = 0.75\textwidth}
\begin{tabular}{||l|l|l|l|r|r||}
\hline \textbf{Dataset} & \textbf{Trait} & \textbf{Sex} & \textbf{Likelihood} & \textbf{Taxa} & \textbf{Phylogram support ($\bm{\pi}$)} \\
\hline \hline
\textit{Mammal} & Body mass & Mixed & Full & 125 & 0.0 \\ \hline
\textit{Mammal} & Body mass & Mixed & REML & 125 & 0.002 \\ \hline
\textit{Mammal} & Body mass & \Female & Full & 46 & 0.848 \\ \hline
\textit{Mammal} & Body mass & \Female & REML & 46 & 0.846 \\ \hline
\textit{Mammal} & Body mass & \Male & Full & 46 & 0.697 \\ \hline
\textit{Mammal} & Body mass & \Male & REML & 46 & 0.684 \\ \hline
\textit{Mammal} & Brain mass & Mixed & Full & 125 & 0.0035 \\ \hline
\textit{Mammal} & Brain mass & Mixed & REML & 125 & 0.0035 \\ \hline
\textit{Mammal} & Brain mass & \Female & Full & 46 & 0.56 \\ \hline
\textit{Mammal} & Brain mass & \Female & REML & 46 & 0.539 \\ \hline
\textit{Mammal} & Brain mass & \Male & Full & 46 & 0.515 \\ \hline
\textit{Mammal} & Brain mass & \Male & REML & 46 & 0.497 \\ \hline
\textit{COMBINE} & Body mass & Mixed & Full & 217 & 0.0 \\ \hline
\textit{COMBINE} & Body mass & Mixed & REML & 217 & 0.0 \\ \hline
\textit{COMBINE} & Brain mass & Mixed & Full & 174 & 0.0015 \\ \hline
\textit{COMBINE} & Brain mass & Mixed & REML & 174 & 0.00649 \\ \hline
\end{tabular}
\end{adjustbox}
\captionof{table}{
\textbf{Summary of empirical datasets.}
Support for the phylogram over the chronogram ($\pi$) for different datasets and traits.
Brain and body mass is either extracted from \citet[\textit{\textit{Mammal}}]{tsuboi_breakdown_2018} or from the \textit{COMBINE} database~\citep{soria_combine_2021}.
For the \citet[\textit{\textit{Mammal}}]{tsuboi_breakdown_2018} dataset, the individuals can also be split into males (\Male) and females (\Female).
Likehood is either computed with a full likelihood or with a restricted maximum likelihood (REML).
The number of taxa is the number of species with data available for the trait.
\label{table:results-empirical}
}
\end{center}

\newpage
\section{Ancestral trait reconstruction}\label{sec:ancestral-trait-reconstruction}
\begin{center}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{A}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_NeutralPhylo}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{B}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_NeutralChrono}
    \end{minipage}\\
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{C}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_SelectionPhylo}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{D}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_SelectionChrono}
    \end{minipage}
    \captionof{figure}{
    \textbf{Phenotypic distance as a function of branch length.}
    Phenoypic distance ($|\Delta z|$) is computed as the absolute difference of the posterior mean trait value between the ancestral node and the descendent node.
    Across all simulated dataset (100 replicates), we computed the mean phenotypic distance (blue dots) and standard deviation (blue horizontal lines) for each branch.
    Simulations under a neutral regime (top row, panels A and B) and under a moving optimum regime (bottom row, panels C and D).
    Reconstruction of ancestral trait on a phylogram (left column, panels A and C) and a chronogram (right column, panels B and D).
    \label{fig:results-distance}
    }
\end{center}
%TC:endignore
\end{document}

### 2025-02-25 manuscript/main.tex
Plot distance phenotype-branch length with color

@@
 Methodologically, such alternative models of evolution raise issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{uyeda_novel_2014, grabowski_cautionary_2023}.
+Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
 Finally, the optimum can also be allowed to change continuously along the tree, which is interpreted as a trait tracking a moving optimum~\citep{hansen_translating_1996, latrille_detecting_2024}.
@@
 We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
-The parameters of simulations were chosen to mimic an empirical dataset of \textit{Mammal}s.
+The parameters of simulations were chosen to mimic an empirical dataset of mammals.
 In summry, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
@@
 \subsection*{Empirical dataset}
-We analyzed a dataset of body and brain masses from \textit{Mammal}s.
+We analyzed a dataset of body and brain masses from mammals.
 The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
@@
 Moreover, for a trait is under selection for which the pace of optimum changes is also tracking the changes in generation time, the support for a phylogram over a chronogram would be misleading~\citep{litsios_effects_2012}.
-Therefore, we argue that while phylograms can offer valuable insights, they should be used with caution and in conjunction with chronograms to provide a more comprehensive understanding of trait evolution.
+Therefore, we argue that while phylograms can offer valuable insights~\citep{wilson_chronogram_2022}, they should be used with caution and in conjunction with chronograms to provide a more comprehensive understanding of trait evolution.
+Contrarily to our cautious note, \citet{wilson_chronogram_2022} suggested that phylograms are more robust for studying discrete character evolution than chronogram.
+This apparent contradiction is due to the different units of measurement used in the studies, as they measure branch lengths in units of morphological distance rather than substitutions per site~\citep{wilson_chronogram_2022}.
+We thus argue that the use of a phylogram, measured in units of substitutions, could also be useful for studying the evolution of discrete characters, but the same caution should be applied.
 
 % Altogether, what we have done and haven't tested and done yet
-Our results suggest that the use of phylograms can be a valuable tool for modeling trait evolution, particularly when the null model is neutral evolution.
+We acknowledge that our study has some limitations.
 First, we simulated only a subset of different regimes of selection, but more complex scenarios could be included for example with with primary and secondary moving optimums~\citep{hansen_three_2024}.
@@
 Second, while the tested alternative models of evolution (multi-rates BM, OU, multi-optium OU) are standard, they do not encompass the full breadth of available methods, which could be expanded~\citep{pennell_geiger_2014, hohna_revbayes_2016}.
-Finally, the dataset of body and brain masses in \textit{Mammal}s may not be representative of all traits or species and even though our analysis showcase the utility of phylograms, more empirical studies are needed to replicate the findings.
+Finally, the dataset of body and brain masses in mammals may not be representative of all traits or species and even though our analysis showcase the utility of phylograms, more empirical studies are needed to replicate the findings.
 
 % Gene expression level and empirical studies.
-One striking example of traits that could (and should) be studied with a phylogram is gene expression level, both for protein concentration and mRNA expression.
+One striking example of traits that could (and should) be empirically studied with a phylogram is gene expression level, both for protein concentration and mRNA expression.
 Indeed, the regime of selection acting on gene expression level is the focus of intense debates~\citep{signor_evolution_2018, price_detecting_2022, bertram_cagee_2023, dimayacyac_evaluating_2023}.
 And typically, dataset on which the Brownian process is favored over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
-More generally, empirical studies plot phenotypic distance as a function of time while claiming that a trait is evolving under drift~\citep{jiang_decoupling_2023}.
-As such, we argue that phenotypic distance should be computed as the function of branch length in unit of substitutions (square root), not time.
-Altogether, the use of a phylogram, or more generally nucleotide divergence could help to disentangle the effect of drift and selection acting on gene expression levels.
-
-% Perspective on the use of phylograms in phylogenetic comparative methods.
-The use of phylograms in phylogenetic comparative methods has been a subject of debate~\citep{wilson_chronogram_2022}.
-The study of \citet{wilson_chronogram_2022} suggested that phylograms are more robust for studying discrete character evolution, contrary to our results suggesting more false positives.
-However, this apparent contradiction is due to the different units of measurement used in the studies, as they measure branch lengths in units of morphological distance rather than substitutions per site.
-We argue that the use of a phylogram, measured in units of substitutions, could also be useful for studying the evolution of discrete characters~\citep{wilson_chronogram_2022}.
-Phylograms provide a valuable tool for modeling trait evolution, especially when the null model is neutral evolution.
-In addition to the neutrality index for a trait~\citep{latrille_detecting_2024}, the toolbox for studying trait evolution could be enriched by the use of phylograms to model trait evolution, particularly when variation within species is available.
-We argue, that the soundness of studying trait evolution on phylograms can be evaluated by the absolute fit of a model to the data~\citep{pennell_model_2015}.
-More generally, genomic information could potentially be seen as a way to disentangle congruence models~\citep{louca_extant_2020}, or as prior for methods that detect shifts in adaptive regimes~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020}.
+Phenotypic distance is also plotted as a function of time while claiming that a trait is evolving under drift~\citep{jiang_decoupling_2023}.
+Instead, we argue that phenotypic distance should be shown as the function of branch length in unit of substitutions (square root), not time.
+More generally, from an empirical perspective, the use of substitutions at neutral loci brings new perspective, particularly when variation within species is also available and the neutrality index for a trait can be computed~\citep{latrille_detecting_2024}.
 
 % Conclusion
-Altogether, we support the use a chronogram when testing for selection acting on a trait.
+Altogether, our study support the use a chronogram when testing for selection acting on a trait.
 The pipeline of hypothesis testing and model comparison is currently well established and should be followed.
@@
 We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis.
-Only then we can we have a bit more support that a trait is evolving under drift if the phylogram has more support than chronogram, but caution is still warranted.
+Only then can we claim that a trait is evolving under drift if the phylogram has more support than chronogram, but caution is still warranted.
 
@@
 From the empirical mammalian dataset (see Methods), we computed an average nucleotide divergence from the root to leaves of $0.18$.
-We scaled parameters in our simulations to fit plausible values for \textit{Mammal}s.
+We scaled parameters in our simulations to fit plausible values for mammals.
 We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per site per generation and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
@@
     Across all simulated dataset (100 replicates), we computed the mean phenotypic distance (blue dots) and standard deviation (blue horizontal lines) for each branch.
-    Simulations under a neutral regime (top row, panels A and B) and under a moving optimum regime (bottom row, panels C and D).
-    Reconstruction of ancestral trait on a phylogram (left column, panels A and C) and a chronogram (right column, panels B and D).
+    Simulations under a neutral regime (yellow, top row, panels A and B) and under a moving optimum regime (bottom row, panels C and D).
+    Reconstruction of ancestral trait on a phylogram (blue, left column, panels A and C) and a chronogram (right column, panels B and D).
     \label{fig:results-distance}

### 2025-02-25 scripts/libraries_plot.py

@@ -23,6 +23,11 @@ YELLOW = "#F0E442"
 CYAN = "#56B4E9"
 WHITE = "#FFFFFF"
 BLACK = "#000000"
+cs_simu_models = {"N": (ORANGE, "Neutral"), "M": (RED, "Multiple optima"), "B": (BLUE, "Moving optimum")}
+
+
+def gr_simu_models(x):
+    return "N" if ("neutral" in x) else ("M" if (("multi" in x) and ("optimum" in x)) else "B")
 
 
 def tex_f(x):
@@ -40,20 +45,11 @@ def tex_f(x):
     return s
 
 
-def colors(x):
-    cs = [RED, BLUE, YELLOW, ORANGE, CYAN, GREEN, PURPLE, BLACK]
-    return [cs[idx % len(cs)] for idx in range(len(x))]
-
-
-def gr_simu_models(x):
-    return "N" if ("neutral" in x) else ("B" if (("multi" in x) and ("optimum" in x)) else "S")
-
-
 def color_simu_models(x_list):
-    cs = {"N": (ORANGE, "Neutral"), "B": (RED, "Multiple optima"), "S": (BLUE, "Moving optimum")}
     grs = [gr_simu_models(xl) for xl in x_list]
-    color_list = [cs[g][0] for g in grs]
-    handles = [Rectangle((0, 0), 1, 1, color=v[0], ec="k", lw=1, label=v[1]) for c, v in cs.items() if c in set(grs)]
+    color_list = [cs_simu_models[g][0] for g in grs]
+    handles = [Rectangle((0, 0), 1, 1, color=v[0], ec="k", lw=1, label=v[1])
+               for c, v in cs_simu_models.items() if c in set(grs)]
     return color_list, handles
 
 
@@ -106,7 +102,7 @@ def vert_boxplot(x_input, y_label, output, yscale="linear", format_label=None, e
     df = pd.DataFrame({"y": np.concatenate([x[k] for k in sorted_x]),
                        "x": np.concatenate([[k] * len(x[k]) for k in sorted_x])})
     sns.violinplot(data=df, x="x", y="y", ax=ax, palette=color_models, log_scale=(yscale == "log"), inner="stick",
-                   cut=0, legend=False, hue="x", linewidth=0.25, linecolor="auto")
+                   cut=0, legend=False, hue="x", linewidth=0.5, linecolor="auto")
     ax.set_xlabel("")
     if "".join(sorted_x).count("Both") > 0:
         handles = []

### 2025-02-25 scripts/plot_trait_distance_RevBayes.py

@@ -4,27 +4,28 @@ import numpy as np
 from collections import defaultdict
 from libraries import *
 import matplotlib.pyplot as plt
+from libraries_plot import cs_simu_models, gr_simu_models
 
 my_dpi = 96
 fontsize = 24
 fontsize_legend = 18
 
 
-def scatter_plot(data, ax: plt.Axes, title: str, x_label: str, y_label: str, average: bool = False):
+def scatter_plot(data, ax: plt.Axes, title: str, x_label: str, y_label: str, average: bool = False, color: str = "black"):
     r2_list = []
     for key, (x, y, y_std) in data.items():
         if y_std is not None:
             # Display the range of the inference nodes, hide the center points
-            ax.errorbar(x, y, yerr=y_std, fmt='none', alpha=0.1, zorder=-1)
+            ax.errorbar(x, y, yerr=y_std, fmt='none', alpha=0.1, zorder=-1, ecolor=color)
         assert len(x) == len(y), "Different number of x and y values"
         # regression line and r-squared
         a, b = np.polyfit(x, y, 1)
         r2 = np.corrcoef(x, y)[0, 1] ** 2
         x_lim = np.array([min(0, min(x)), max(x)])
         if average:
-            ax.scatter(x, y, alpha=0.25)
+            ax.scatter(x, y, alpha=0.25, color=color)
         else:
-            ax.scatter(x, y, label=f"{key} (n = {len(x)})")
+            ax.scatter(x, y, label=f"{key} (n = {len(x)})", color=color)
         ax.plot(x_lim, a * x_lim + b, color="black", linestyle="--", label=f"Linear regression (R² = {r2:.2f})")
         r2_list.append(r2)
     # ax.axhline(0, color="black", linestyle="--")
@@ -68,21 +69,15 @@ def main(input_distance_tree: str, input_annotated_tree_list: list[str], output_
             trait_parent = float(getattr(n_t.up, "Phenotype_mean"))
             dict_y[n_t.name].append(np.abs(trait - trait_parent))
 
-
-
     x_label = "$\\sqrt{" + ("d" if "Phylo" in os.path.basename(input_distance_tree) else "T") + "}$"
     y_label = "|Δz|"
     fig, ax = plt.subplots(1, 1, figsize=(640 / my_dpi, 480 / my_dpi), dpi=my_dpi)
     x_mean = {k: np.mean(v) for k, v in dict_x.items()}
[…]

### 2025-02-28 manuscript/main.tex
Update manuscript

@@
 %TC:ignore
+
 \maketitle
@@
 This species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
-    Theoretically, changes for a neutral trait depend on the number of generations passed, which in turn correlates with time, hence the use of chronogram.
+Indeed, the time that has passed correlates with the changes for a neutral trait, through the direct proxy of the number of generations that occurred.
 But, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases.
-    Instead, for a phylogram the branch lengths represent sequence divergence (i.e. the number of substitutions), which absorb the effect of changing generation time and mutation rates.
-    In this study, we first show using simulations that for a trait evolving neutrally the fit of a Brownian process has more support on a phylogram than on a chronogram.
+Instead, for a phylogram, the branch lengths represent sequence divergence (i.e.
+the number of substitutions) which absorb the effect of changing generation time.
+Passed on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a Brownian process has more support on a phylogram than on a chronogram.
 However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions.
@@
 
-\keywords{Trait evolution \and Phylogenetics \and Phylogenetic Comparative Methods \and Chronogram \and Sequence divergence \and Neutral evolution}
+\keywords{Comparative studies \and Morphological evolution \and Genetic drift \and Macroevolution \and Chronogram \and Phylogram}
 
@@
 % The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
-Across species, distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed variation.
-Such regimes of evolution are typically compared using phylogenetic comparative methods, where a continuous traits is modeled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
+By observing the changes in trait along a lineage, it is possible to speculate on the regime of selection acting on it.
+But without selection, a trait might still changes due to random genetic drift.
+As such distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
+At a larger scale, from the observed variations of traits across species, such regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
 For example, to model neutral evolution, the mean trait value is said to follow a Brownian Motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
@@
 One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
-This modeling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a fluctuating optimum.
+This modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a fluctuating optimum.
 
@@
 % The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
-One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}).
+One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
 Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
@@
 But, since species generation time might vary along the phylogenetic tree, modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
-Instead, on a phylogram, the branch lengths can represent another quantity than time.
-For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction~\citep{wilson_chronogram_2022}.
+Instead, on a phylogram, the branch lengths can represent another quantity than time, which can be used as the backbone to model trait evolution.
+For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}.
 From a genomic perspective, phylogram can also be obtained in unit of nucleotide divergence, which is equivalent to measuring the number of nucleotide substitutions occurring along a branch.
-Such a phylogram in unit of nucleotide divergence obtained from neutrally evolving genomic loci would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
+Such a phylogram in unit of nucleotide divergence, if obtained from neutrally evolving genomic loci, would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
 Conversely, for a trait evolving under selection, using a chronogram would be more accurate since changes in the optimum fitness peak is extrinsic to the species and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
@@
     \caption{
-        Panel~A: Across many species, the evolution of continuous traits can be modeled as a stochastic process evolving along the branches of a phylogenetic tree.
+        Panel~A: Across many species, the evolution of continuous traits can be modelled as a stochastic process evolving along the branches of a phylogenetic tree.
         The branches of such phylogenetic tree can be measured in either time (chronogram) or number of substitutions (phylogram).
@@
 % Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
-In this study, we test if it is more accurate to use a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram as the underlying tree to model neutral trait changes, by evaluating by the fit of a BM on both trees (Fig.~\ref{fig:methods}B).
+In this study, we test if it is more accurate to use a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram as the underlying tree to model neutral trait changes (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
 In other words, we seek to test if change of a neutral trait is better predicted a phylogram.
@@
 \subsection*{Simulations along a phylogenetic tree}
-We performed simulations under different selective regimes (neutral, multiple optima, moving optimum).
+We performed simulations under different selective regimes (neutral, moving optimum, multiple optima).
 Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
@@
 The parameters of simulations were chosen to mimic an empirical dataset of mammals.
-In summry, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
+In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
 Parents were selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
@@
 The phylogram is obtained from the same simulation setting on a set of independent 30.000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
-The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood smoothing parameter of 1) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
+The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
 
 \subsection*{Empirical dataset}
-We analyzed a dataset of body and brain masses from mammals.
+We analysed a dataset of body and brain masses from mammals.
 The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
@@
 More specifically, the phylogram in unit of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
-The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood smoothing parameter of 1) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
+The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
 
@@
 The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
-As in the previous section, a Brownian process is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length.
+As in the previous section, a Brownian process is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
 Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
 Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
-The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favored over the chronogram.
+The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favoured over the chronogram.
 
@@
 
-\section*{Results}
-
-\subsection*{Support for a phylogram over a chronogram}
-% Simulations of a trait evolving neutrally versus under a moving optimum.
-We developed a Bayesian model to estimate the support of phylogram over a chronogram for a trait evolving under a Brownian Motion (BM) process (see Methods, BM section).
-To test this model, we simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) evolving under multiple optima, and 3) evolving under a moving optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
-As a control experiment we simulated with a constant generation time, mutation rate (per generation) and effective population size ($\Ne$).
-In such a case, the support is the same for both a phylogram and a chronogram (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
-Next, we simulated changing generation time, mutation rate and $\Ne$ along the phylogenetic tree mimicking a mammalian range of changes.
-In this case, the BM was better fitted on a phylogram than on a chronogram for a neutral trait (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
-Conversely, the fit of a BM on a chronogram was better than on a phylogram for a trait under selection, with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
-On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
-When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
-For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
-On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
-We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).
-
 %TC:ignore
-\begin{figure*}[!ht]
+\begin{figure*}[!htb]
     \centering
@@
         Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
-        Panel~B: Drift is modeled by the resampling of parents.
-        Selection is modeled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree.
+        Panel~B: Neutral evolution is modelled as a constant fitness regardless of phenotypic value.
+        Selection is modelled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree.
         Panel~C: Example simulation of a trait evolving along a phylogenetic tree.
-        Panels D \& E: Violinplot of posterior probability of the phylogram being favored over the chronogram for the different simulated regime of selection.
-        Horizontal lines inside the violin show the result of each replicate simulation.
-        Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow), under a moving optimum (blue), and under multiple optima (red).
+        Panels D \& E: Violinplot of posterior probability of the phylogram being favoured over the chronogram ($\PhyloSupport$) across 100 replicate simulations for the different simulated regime of evolution.
+        Horizontal lines inside violins are the estimated $\PhyloSupport$ of each replicate simulation.
+        Regime of evolution: neutral regime (yellow), under a moving optimum (blue), and under multiple optima (red).
         Results of simulations with constant generation time (Panel~D) and with fluctuating generation time, mutation rate and effective population size (Panel~E).
@@
 
-\subsection*{Trait evolution on a phylogram and on a chronogram}
-% Mis-classification of trait evolution on a phylogram.
-Beside support for the underlying tree, ancestral state reconstruction can also be improved by the use of a chronogram or a phylogram.
-For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
-In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$)
-than a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).
-Additionally, instead of testing the fit of a BM with a single rate, we also fitted a multi-rate BM process (see Methods).
-When fitting a multi-rate BM process, the estimated variance of rate parameters ($v$) was lower on a phylogram than on a chronogram for a trait evolving neutrally (Fig.~\ref{fig:results-alternative}A, yellow violins, Wilcoxon paired test with $p_{\\text{value}}=8.8\times 10^{-14}$), and the number of rate shifts ($n$) was reflecting the prior on the phylogram while not on the chronogram, showing a bias (Fig.~\ref{fig:results-alternative}B, yellow violins).
-Conversely, for a trait under a moving optimum, $v$ was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}A, blue violins, Wilcoxon paired test with  $p_{\\text{value}}=8.9\times 10^{-15}$), and $n$ was reflecting the prior on the chronogram while not on the phylogram, showing a bias (Fig.~\ref{fig:results-alternative}B, blue violins).
-Altogether, fitting a BM on a phylogram is more accurate for a trait evolving neutrally, but results in less accurate estimates for a trait under selection.
+\section*{Results}
 
-% Mainly testing BM versus OU
-Beside testing the fit of a BM, we tested alternative Ornstein-Uhlenbeck (OU) models: with a single optimum and multi-optima (see Methods).
-First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\\text{value}}=3.9\times 10^{-18}$).
-Worst still, even for a trait evolving neutrally, $p_{\\text{OU}}$ was still higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, yellow violins, Wilcoxon paired test with $p_{\\text{value}}=8.0\times 10^{-8}$).
-When fitting a multi-OU process on a trait evolving neutrally, the number of optimum shifts ($m$) was reflecting the prior on the phylogram while not on the chronogram (Fig.~\ref{fig:results-alternative}B, yellow violins).
-Conversely, for a trait under a moving optimum, ($m$) was reflecting the prior on the chronogram while not on the phylogram (Fig.~\ref{fig:results-alternative}A, blue violins).
-Altogether, fitting an OU process on a phylogram is more accurate for a trait evolving under selection, but results in less accurate estimates for a trait evolving neutrally.
+\subsection*{Support for a phylogram over a chronogram}
+% Simulations of a trait evolving neutrally versus under a moving optimum.
+We developed a Bayesian model to estimate the support of phylogram over a chronogram for a trait evolving under a Brownian Motion (BM) process, denoted as $\PhyloSupport$, which range from 0 (full support for the chronogram) to 1.0  (full support for the phylogram), for more details see Methods (BM section).
+To test this model, we simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum and 3) under multiple optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
+As a control experiment we simulated with a constant generation time, mutation rate (per generation) and effective population size ($\Ne$).
+In such a case, the support is the same for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
+Next, we simulated changing generation time, mutation rate and $\Ne$ along the phylogenetic tree, mimicking a mammalian range of changes.
+In this case, the BM was better fitted on a phylogram than on a chronogram for a neutral trait (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
+Conversely, the fit of a BM on a chronogram was better than on a phylogram for a trait under selection, with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
+On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
+When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
+For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
+On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
+We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).
 
 %TC:ignore
-\begin{figure*}[!ht]
+\begin{figure*}[!htb]
     \centering
@@
         Violinplot of posterior parameter estimates for different models of trait evolution and different simulated regime of selection.
-        Horizontale lines inside the violin show the result of each replicate simulation.
+        Horizontal lines inside the violin show the result of each replicate simulation.
         Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow) and under a moving optimum (blue).
-        Wilcoxson rank-test are performed between the paired estimates on the phylogram and the chronogram.
+        Wilcoxon rank-test are performed between the paired estimates on the phylogram and the chronogram.
         Panel~A \& B: Fit of a relaxed Brownian Motion (BM) with multiple rate parameters, using either a phylogram or a chronogram.
@@
 
+\subsection*{Trait evolution on a phylogram and on a chronogram}
+% Mis-classification of trait evolution on a phylogram.
+Beside support for the underlying tree, ancestral state reconstruction can also be improved by the use of a chronogram or a phylogram.
+For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
+In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$) than on a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).
+Additionally, instead of testing the fit of a BM with a single rate, we also fitted a multi-rate BM process (see Methods).
+When fitting a multi-rate BM process, the estimated variance of rate parameters ($v$) was lower on a phylogram than on a chronogram for a trait evolving neutrally (Fig.~\ref{fig:results-alternative}A, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.8\times 10^{-14}$), and the number of rate shifts ($n$) was reflecting the prior on the phylogram while not on the chronogram, showing a bias (Fig.~\ref{fig:results-alternative}B, yellow violins).
+Conversely, for a trait under a moving optimum, $v$ was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}A, blue violins, Wilcoxon paired test with  $p_{\text{value}}=8.9\times 10^{-15}$), and $n$ was reflecting the prior on the chronogram while not on the phylogram, showing a bias (Fig.~\ref{fig:results-alternative}B, blue violins).
+Altogether, fitting a BM on a phylogram is more accurate for a trait evolving neutrally, but results in less accurate estimates for a trait under selection.
+
+% Mainly testing BM versus OU
+Beside testing the fit of a BM, we tested alternative Ornstein-Uhlenbeck (OU) models: with a single optimum and with multiple optima (see Methods).
+First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\text{value}}=3.9\times 10^{-18}$).
+Worst still, even for a trait evolving neutrally, $p_{\text{OU}}$ was still higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.0\times 10^{-8}$).
+When fitting a multi-OU process on a trait evolving neutrally, the number of optimum shifts ($m$) was reflecting the prior on the phylogram while not on the chronogram (Fig.~\ref{fig:results-alternative}B, yellow violins).
+Conversely, for a trait under a moving optimum, ($m$) was reflecting the prior on the chronogram while not on the phylogram (Fig.~\ref{fig:results-alternative}A, blue violins).
+Altogether, fitting an OU process on a phylogram results in biases, for a trait under selection but even for a trait under neutral evolution.
+
 \section*{Discussion}
 % Context and rationale
-In phylogenetic comparative methods, the evolution of continuous traits is typically modeled as a stochastic process running along the branches of a phylogenetic tree.
+In phylogenetic comparative methods, the evolution of continuous traits is typically modelled as a stochastic process running along the branches of a phylogenetic tree.
 Such branches are measured in either time (tree is a chronogram) or in number of nucleotide substitutions (tree is a phylogram).
 Typically, the null model of neutral trait evolution (i.e.~drift) is though as a Brownian process running on a chronogram~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
-As a result, dataset on which the Brownian process is favored over alternative models are often interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
+As a result, traits on which the Brownian process is favoured over alternative models are interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
 However, we argue instead that claims for a trait evolving under drift should be tested more thoroughly by comparing the fit of the Brownian process under both a phylogram and a chronogram as the underlying tree.
@@
 Conversely, for simulation under a moving optimum, the fit of a Brownian process on a chronogram was better than on a phylogram for a trait.
-Applied to different empirical dataset of body and brain masses in mammals, we showed that the chronogram was favored over the phylogram, ruling out the neutral model of evolution for these trait~\citep{latrille_detecting_2024}.
+Applied to different empirical dataset of body and brain masses in mammals, we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these trait~\citep{latrille_detecting_2024}.
 
@@
 Moreover, if population structure changes of the probability of fixation of neutral mutations (that would also impact a neutral trait), the phylogram would absorb automatically these changes.
-This argument is not only verbal since we tested for it in our simulation by modelling changes in both $\Ne$ and generation time, and the phylogram was favored over the chronogram for a trait evolving neutrally.
+This argument is not only verbal since we tested for it in our simulation by modelling changes in both $\Ne$ and generation time, and the phylogram was favoured over the chronogram for a trait evolving neutrally.
 Second, aside from absorbing of changes in generation time and $\Ne$, the use of phylograms also absorbs changes in mutation rate (per generation) under certain assumptions.
@@
 These examples show that while phylograms are useful to neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
-In our simulations, we found that generally using a phylogram tend to favor the OU process over the Brownian Motion (BM) model.
-This suggests that trees in unit of nucleotide divergence may not always provide a reliable framework for modeling trait evolution, especially when selection is involved.
-As such, we recommend continuing to use chronograms for modeling selection acting on a trait, as they generally provide more accurate and reliable results.
+In our simulations, we found that generally using a phylogram tend to favour the OU process over the Brownian Motion (BM) model.
+This suggests that trees in unit of nucleotide divergence may not always provide a reliable framework for modelling trait evolution, especially when selection is involved.
+As such, we recommend continuing to use chronograms for modelling selection acting on a trait, as they generally provide more accurate and reliable results.
 Moreover, for a trait is under selection for which the pace of optimum changes is also tracking the changes in generation time, the support for a phylogram over a chronogram would be misleading~\citep{litsios_effects_2012}.
@@
 Such constant architecture result in saturation of the phenotype on long time scales, with a phenomenon similar to nucleotide saturation~\citep{latrille_detecting_2024}.
-Such saturation might be one of the reason for a bias in favor of the Ornstein-Uhlenbeck (OU) process over the Brownian Motion (BM) model that we observed in our simulations.
+Such saturation might be one of the reason for a bias in favour of the Ornstein-Uhlenbeck (OU) process over the Brownian Motion (BM) model that we observed in our simulations.
 A more realistic genetic architecture should include pleiotropy, epistasis and linkage disequilibrium, while here we only considered non-linked loci contributing additively to a single trait.
-Second, while the tested alternative models of evolution (multi-rates BM, OU, multi-optium OU) are standard, they do not encompass the full breadth of available methods, which could be expanded~\citep{pennell_geiger_2014, hohna_revbayes_2016}.
+Second, while the tested alternative models of evolution (multi-rates BM, OU, multi-optimum OU) are standard, they do not encompass the full breadth of available methods, which could be expanded~\citep{pennell_geiger_2014, hohna_revbayes_2016}.
 Finally, the dataset of body and brain masses in mammals may not be representative of all traits or species and even though our analysis showcase the utility of phylograms, more empirical studies are needed to replicate the findings.
@@
 Indeed, the regime of selection acting on gene expression level is the focus of intense debates~\citep{signor_evolution_2018, price_detecting_2022, bertram_cagee_2023, dimayacyac_evaluating_2023}.
-And typically, dataset on which the Brownian process is favored over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
+And typically, dataset on which the Brownian process is favoured over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
 Phenotypic distance is also plotted as a function of time while claiming that a trait is evolving under drift~\citep{jiang_decoupling_2023}.
@@
 The pipeline of hypothesis testing and model comparison is currently well established and should be followed.
-However, when the analysis concludes that the BM is favored, then the support of the phylogram over the chronogram should be tested additionally.
+However, when the analysis concludes that the BM is favoured, then the support of the phylogram over the chronogram should be tested additionally.
 We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis.
@@
 
-The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modeled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
-These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0043\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0043\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0043 \Multiply \sqrt {13,500} = 0.5$.
+The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modelled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
+These processes are parametrized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0043\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0043\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0043 \Multiply \sqrt {13,500} = 0.5$.
 In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $0.5$.
@@
 For the \citet[\textit{\textit{Mammal}}]{tsuboi_breakdown_2018} dataset, the individuals can also be split into males (\Male) and females (\Female).
-Likehood is either computed with a full likelihood or with a restricted maximum likelihood (REML).
+Likelihood is either computed with a full likelihood or with a restricted maximum likelihood (REML).
 The number of taxa is the number of species with data available for the trait.
@@
 %TC:endignore
+
+\newpage
+
+\begin{center}
+    Cover Letter
+\end{center}
+
+Dear Editor,
+
+We are pleased to submit our manuscript entitled ”Phylogram for testing neutral evolution of a trait, chronogram to model selection” for consideration for publication in Evolution Letters.
+
+In this study, we revisit a puzzling question of phylogenetic comparative methods: does the time that have passed since the species diverged explain differences in traits?
+Or is it the genetic differences between species that explains best the differences?
+While the de facto standard in PCM is to use the time to explain the differences in traits, this reasoning is not always justified.
+Indeed, if the trait is evolving neutrally, we argue that the genetic differences between species instead of time should explain best the differences in traits.
+As a result, in this study, we show and argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.
+
+We believe that our manuscript will be of great interest to the readers of Evolution Letters as it offers a novel perspective on trait evolution and provides practical guidelines for researchers in the field.
+We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.
+
+With best regards,
+
+Thibault Latrille, Thibault Gaboriau and Nicolas Salamin
+
 \end{document}

### 2025-03-17 manuscript/main.tex
Updated manuscript with Théo. G comments

@@
 
-\renewcommand{\baselinestretch}{1.5}
+\renewcommand{\baselinestretch}{1.4}
 \renewcommand{\arraystretch}{1.2}
@@
 \begin{abstract}
-The evolution of traits across different species is driven by a combination of selective and neutral processes, while their relative importance is subject to numerous debates.
-To disentangle them, different scenarios of evolution are modelled and compared.
-Typically, a neutrally evolving trait is modelled as changing randomly along the different branches of the species tree.
-This species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
-Indeed, the time that has passed correlates with the changes for a neutral trait, through the direct proxy of the number of generations that occurred.
-But, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases.
+At the species level, the evolution of traits is driven by a combination of selective and neutral forces.
+To disentangle these processes, different scenarios of evolution are modelled and compared.
+For traits under selection, the species is seen as tracking a trait optimum, and such optimum can change along the species tree.
+On the other hand, neutral evolution is modelled as the trait changing randomly along the different branches of the species tree.
+But regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
+Rationally, the time that has passed correlates with trait changes, even under a neutral regime, through the direct proxy of the number of generations that occurred.
+However, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases.
 Instead, for a phylogram, the branch lengths represent sequence divergence (i.e.
 the number of substitutions) which absorb the effect of changing generation time.
-Passed on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a Brownian process has more support on a phylogram than on a chronogram.
+Passed on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a random evolution has more support on a phylogram than on a chronogram.
 However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions.
@@
 
+\newpage
 %TC:endignore
@@
 By observing the changes in trait along a lineage, it is possible to speculate on the regime of selection acting on it.
-But without selection, a trait might still changes due to random genetic drift.
+But even without selection, a trait might still changes due to random genetic drift.
 As such distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
 At a larger scale, from the observed variations of traits across species, such regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
-For example, to model neutral evolution, the mean trait value is said to follow a Brownian Motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
+For example, to model neutral evolution, the mean trait value is said to follow a Brownian motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
 Or in other words, for each  branch of the tree the value at descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
@@
 One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
-This modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a fluctuating optimum.
+This modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a continuously moving optimum.
 
@@
 Such a phylogram in unit of nucleotide divergence, if obtained from neutrally evolving genomic loci, would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
-Conversely, for a trait evolving under selection, using a chronogram would be more accurate since changes in the optimum fitness peak is extrinsic to the species and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
+Conversely, for a trait evolving under selection at the species level, using a chronogram would be more accurate to model mean trait changes since shifts in the optimum fitness peak is extrinsic to the species, and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
 % Finally, from a statistical perspective, using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g.
@@
 % Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
-In this study, we test if it is more accurate to use a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram as the underlying tree to model neutral trait changes (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
-In other words, we seek to test if change of a neutral trait is better predicted a phylogram.
+In this study, we test if using a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram can allow to assess the regime of evolution of a trait (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
+More precisely, we seek to test if change of a neutral trait is better predicted a phylogram.
 Conversely, we also seek to test if changes of a trait under a moving optimum is better predicted by a chronogram.
-Altogether, by using genomic information we seek to provide additional approaches to model trait evolution.
+As such, by using genomic information we seek to provide additional approaches to model trait evolution, particularly to disentangle selection and neutral evolution.
 
 \section*{Methods}
+In order to assess whether a phylogram is better suited to model neutral evolution we first performed simulations of a trait evolving along a phylogenetic tree under different regimes of selection (neutral, moving optimum, multiple optima).
+Second, we gathered an empirical dataset of body and brain masses from mammals, including both a chronogram and a phylogram (in unit of nucleotide divergence from neutral loci).
+Third, we fitted a single rate Brownian motion (BM) to the simulated data, and we compared the fit of the BM on a phylogram versus a chronogram.
+Forth, based in the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated and empirical datasets.
+Finally, we fitted several models of trait evolution to the simulated and empirical datasets: a multi-rate BM, an Ornstein-Uhlenbeck (OU) process, and a relaxed OU with multiple optima.
+
 \subsection*{Simulations along a phylogenetic tree}
@@
 Flattening the fitness landscape result in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
-Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
+Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
 Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch, if a switch occurs the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).
@@
 As a control, we performed simulations with constant mutation rate, generation times and effective population sizes ($\Ne$).
-Alternatively, we performed simulations with fluctuating mutation rate, generation times and $\Ne$, were we used a Brownian process to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).
+Alternatively, we performed simulations with fluctuating mutation rate, generation times and $\Ne$, were we used a BM to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).
 
@@
 
-\subsection*{Brownian Motion (BM)}
-We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian Motion (BM) on continuous traits evolution evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
+\subsection*{Brownian motion (BM)}
+On the simulated dataset, we fitted a single rate BM using either a phylogram or a chronogram as the underlying tree.
+We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian motion (BM) on continuous traits evolution evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
 The data consist of the mean trait value for each extant species, and the tree topology is fixed.
@@
 Importantly, the data and the tree topology are the same in both analyses, only the branch lengths are different.
+We expect for trait under neutral evolution more accurate trait reconstruction on a phylogram, and conversely for trait under selection (moving optimum) more accurate trait reconstruction on a chronogram.
 
-Additionally to a single rate BM, we fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree.
+\subsection*{BM with a switch}
+To test the support of a phylogram over a chronogram for the simulated and empirical datasets, we implemented a model based on a single rate BM.
+Such model was implemented in \textit{RevBayes} and contains a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
+Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
+If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
+The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
+As in the previous section, a BM is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
+Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
+Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
+The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favoured over the chronogram.
+We expect $\PhyloSupport = 1$ for trait under neutral evolution, and $\PhyloSupport = 0$ for trait under selection (moving optimum).
+
+\subsection*{Alternatives to single rate BM}
+Additionally to a single rate BM, we fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
 The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
@@
 
-Finally, we implemented a model in \textit{RevBayes} containing a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
-Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
-If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
-The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
-As in the previous section, a Brownian process is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
-Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
-Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
-The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favoured over the chronogram.
-
-\subsection*{Ornstein-Uhlenbeck (OU)}
 % Simple OU model versus a brownian process.
-First, we compared the fit of an Ornstein-Uhlenbeck (OU) process (eq.~\ref{eq:ou-process}) to a BM~\citep{hansen_stabilizing_1997,butler_phylogenetic_2004}, using either a phylogram or a chronogram as the underlying tree.
+Also, we compared the fit of an Ornstein-Uhlenbeck (OU) process (eq.~\ref{eq:ou-process}) to a BM~\citep{hansen_stabilizing_1997,butler_phylogenetic_2004}, using either a phylogram or a chronogram as the underlying tree.
 % Ornstein-Uhlenbeck model equations
@@
 In the OU process, the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, and $\theta$ is the optimum value.
-The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the Brownian process, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
+The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the BM, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
 % alpha ~ dnReversibleJumpMixture(0.0, dnExponential( abs(root_age / 2.0 / ln(2.0)) ), 0.5)
@@
 Thus, when $\alpha = 0$ the OU process is equivalent to a BM.
-The support for the OU model is estimated as the posterior mean of the switch variable, or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).
+The support for the OU model is estimated as the posterior mean of the switch variable ($p_{\text{OU}}$), or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).
 
 % Relaxed OU model, how many optimums?
-Second, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
+Finally, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
 The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1000 gen., running of 5000 gen., 2 chains).
@@
 % Simulations of a trait evolving neutrally versus under a moving optimum.
-We developed a Bayesian model to estimate the support of phylogram over a chronogram for a trait evolving under a Brownian Motion (BM) process, denoted as $\PhyloSupport$, which range from 0 (full support for the chronogram) to 1.0  (full support for the phylogram), for more details see Methods (BM section).
-To test this model, we simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum and 3) under multiple optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
-As a control experiment we simulated with a constant generation time, mutation rate (per generation) and effective population size ($\Ne$).
-In such a case, the support is the same for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
-Next, we simulated changing generation time, mutation rate and $\Ne$ along the phylogenetic tree, mimicking a mammalian range of changes.
-In this case, the BM was better fitted on a phylogram than on a chronogram for a neutral trait (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
-Conversely, the fit of a BM on a chronogram was better than on a phylogram for a trait under selection, with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
+We first tested the hypothesis that a phylogram is better suited to model neutral evolution than a chronogram.
+We simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum and 3) under multiple optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
+We tested for the support of phylogram over a chronogram under the assumption that changes follow a Brownian motion (BM) process, we developed as statistic denoted as $\PhyloSupport$, which range from 0 (full support for the chronogram) to 1.0 (full support for the phylogram).
+For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
+Next, we simulated changing generation time, mutation rate (per generation) and effective population size ($\Ne$) along the phylogenetic tree, mimicking a mammalian range of changes.
+In this case, for a neutral trait, the BM was better fitted on a phylogram than on a chronogram (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
+Conversely, for a trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
+
 On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
@@
         Wilcoxon rank-test are performed between the paired estimates on the phylogram and the chronogram.
-        Panel~A \& B: Fit of a relaxed Brownian Motion (BM) with multiple rate parameters, using either a phylogram or a chronogram.
+        Panel~A \& B: Fit of a relaxed Brownian motion (BM) with multiple rate parameters, using either a phylogram or a chronogram.
         Posterior estimate for the variance in rate parameters (Panel~A) and number of rate changes (Panel~B).
@@
 % Mis-classification of trait evolution on a phylogram.
-Beside support for the underlying tree, ancestral state reconstruction can also be improved by the use of a chronogram or a phylogram.
+Beside support for the underlying tree, we also assessed the accuracy of ancestral trait reconstruction on a phylogram versus a chronogram.
 For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
@@
 % Mainly testing BM versus OU
-Beside testing the fit of a BM, we tested alternative Ornstein-Uhlenbeck (OU) models: with a single optimum and with multiple optima (see Methods).
+Moreover, we tested alternative to the BM process with Ornstein-Uhlenbeck (OU) models: with a single optimum and with multiple optima (see Methods).
 First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\text{value}}=3.9\times 10^{-18}$).
@@
 In phylogenetic comparative methods, the evolution of continuous traits is typically modelled as a stochastic process running along the branches of a phylogenetic tree.
-Such branches are measured in either time (tree is a chronogram) or in number of nucleotide substitutions (tree is a phylogram).
-Typically, the null model of neutral trait evolution (i.e.~drift) is though as a Brownian process running on a chronogram~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
-As a result, traits on which the Brownian process is favoured over alternative models are interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
-However, we argue instead that claims for a trait evolving under drift should be tested more thoroughly by comparing the fit of the Brownian process under both a phylogram and a chronogram as the underlying tree.
-Indeed, for a trait evolving neutrally, changes of generation time along the tree would be absorbed by the phylogram while not by the chronogram~\citep{latrille_detecting_2024}, so we argue that the chronogram should not be used as the backbone tree in this case.
-Moreover, using phylogram would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks) to estimate divergence times of chronograms.
-
-% Our main result
-% Claiming that a trait is evolving under drift should be tested by comparing the fit under a phylogram and under a chronogram.
-We developed a Bayesian model implemented in \textit{RevBayes}~\citep{hohna_revbayes_2016} to indicate whether a Brownian process is fitted better on a phylogram or a chronogram ($\PhyloSupport$), given data at the tip of the tree.
-For simulations of a trait evolving neutrally, we showed that the fit of a Brownian process supports a phylogram rather than a chronogram.
-Moreover, for the neutral trait, both ancestral state reconstruction and estimation for the number of rate changes for a multi-rate BM was also more accurate on a phylogram.
-Conversely, for simulation under a moving optimum, the fit of a Brownian process on a chronogram was better than on a phylogram for a trait.
-Applied to different empirical dataset of body and brain masses in mammals, we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these trait~\citep{latrille_detecting_2024}.
+The null model of neutral trait evolution, genetic drift, is though as a Brownian motion (BM) running on the tree~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
+As a result, traits on which the BM is favoured over alternative models are sometimes interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
+However, the BM is not only the null model of neutral evolution, but can also model selection, with a trait tracking a moving optimum, where the optimum itself is following a BM~\citep{hansen_translating_1996, latrille_detecting_2024}.
+Such identifiability is problematic, but we here argue that the underlying backbone tree to model trait evolution can help to disentangle neutral evolution from selection.
+The standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are measured in time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}, while phylograms have been generally overlooked, where the branch lengths are measured in number of substitutions instead.
+Such overlook is rational and justified, since for simulated trait under selection we assessed that the BM has better fit on a chronogram than on a phylogram.
+However, we here argue that to model a trait evolving neutrally, the backbone tree should not be a chronogram but instead a phylogram.
+Such phylogram would better represent the number of generations that occurred, and in turn should explain best the differences in traits between species~\citep{latrille_detecting_2024}
+In practice, we show that for simulations of a trait evolving neutrally, the fit of a BM supports a phylogram rather than a chronogram.
+Additionally, for the neutral trait, both ancestral state reconstruction and estimation for the number of rate changes for a multi-rate BM was also more accurate on a phylogram.
+Altogether, the combined use of a phylogram and a chronogram can help support the hypothesis of neutral evolution, or instead rule it out.
+% Thus, we argue that claims for a trait evolving under drift should be tested more thoroughly by comparing the fit of the BM under both a phylogram and a chronogram as the underlying tree.
 
 % When does it apply and why is it important
+Applied to different empirical dataset of body and brain masses in mammals, we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these trait~\citep{latrille_detecting_2024}.
 The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree.
@@
 First, normalization by nucleotide divergence also accounts theoretically for variations in effective population size ($\Ne$) since the substitution rate of neutral mutations is equal to the mutation rate, such that $\Ne$ has no effect on the rate of neutral evolution~\citep{kimura_evolutionary_1968, ohta_population_1972}.
-Moreover, if population structure changes of the probability of fixation of neutral mutations (that would also impact a neutral trait), the phylogram would absorb automatically these changes.
+Moreover, if population structure impacts the probability of fixation for neutral mutations (that would also impact a neutral trait), the phylogram would absorb automatically these changes.
 This argument is not only verbal since we tested for it in our simulation by modelling changes in both $\Ne$ and generation time, and the phylogram was favoured over the chronogram for a trait evolving neutrally.
@@
 However, if the mutation rate would only changes for a subset of the genome that is encoding the trait, the phylogram would not absorb these changes in mutation rate.
+Finally, because chronogram and usually derived from chronogram by calibrating the tree with molecular clocks, using phylogram has the additional advantage of removing model assumptions required to calibrate node ages~\citep{litsios_effects_2012}.
 
@@
 These examples show that while phylograms are useful to neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
-In our simulations, we found that generally using a phylogram tend to favour the OU process over the Brownian Motion (BM) model.
+In our simulations, we found that generally using a phylogram tend to favour the OU process over the Brownian motion (BM) model.
 This suggests that trees in unit of nucleotide divergence may not always provide a reliable framework for modelling trait evolution, especially when selection is involved.
@@
 % Altogether, what we have done and haven't tested and done yet
-We acknowledge that our study has some limitations.
+Caution is indeed warranted, as there are several limitations to our approach.
 First, we simulated only a subset of different regimes of selection, but more complex scenarios could be included for example with with primary and secondary moving optimums~\citep{hansen_three_2024}.
@@
 Such constant architecture result in saturation of the phenotype on long time scales, with a phenomenon similar to nucleotide saturation~\citep{latrille_detecting_2024}.
-Such saturation might be one of the reason for a bias in favour of the Ornstein-Uhlenbeck (OU) process over the Brownian Motion (BM) model that we observed in our simulations.
+Such saturation might be one of the reason for a bias in favour of the Ornstein-Uhlenbeck (OU) process over the Brownian motion (BM) model that we observed in our simulations.
 A more realistic genetic architecture should include pleiotropy, epistasis and linkage disequilibrium, while here we only considered non-linked loci contributing additively to a single trait.
@@
 Indeed, the regime of selection acting on gene expression level is the focus of intense debates~\citep{signor_evolution_2018, price_detecting_2022, bertram_cagee_2023, dimayacyac_evaluating_2023}.
-And typically, dataset on which the Brownian process is favoured over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
+And typically, dataset on which the BM is favoured over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
 Phenotypic distance is also plotted as a function of time while claiming that a trait is evolving under drift~\citep{jiang_decoupling_2023}.
@@
 However, when the analysis concludes that the BM is favoured, then the support of the phylogram over the chronogram should be tested additionally.
-We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis.
-Only then can we claim that a trait is evolving under drift if the phylogram has more support than chronogram, but caution is still warranted.
+We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis, which indicate whether a BM support better a phylogram or a chronogram ($0 \leq \PhyloSupport \leq 1$), given data at the tip of the tree and both trees with the same topology.
+Only then can we claim that a trait is evolving under drift if the phylogram has more support than chronogram ($\PhyloSupport$ close to $1$), but caution is still warranted.
 
 %TC:ignore
-\printbibliography
 
@@
 Snakemake pipeline, simulator, analysis scripts and \textit{RevBayes} scripts and documentation are available in the repository to replicate the study.
+
+\printbibliography
 
@@
 
-The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modelled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
+The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modelled by a Brownian motion (BM) on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
 These processes are parametrized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0043\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0043\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0043 \Multiply \sqrt {13,500} = 0.5$.
 In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $0.5$.
-An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
+An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric BM to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
 The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
@@
 Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
-Moving optimum selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
+Moving optimum selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric BM~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
 Multiple optima were implemented by allowing the optimum phenotype to shift along a branch with a probability of $10^{-5}$, the shift in optimum being then drawn from a exponential distribution with rate parameter $0.1$, meaning an average of $10$ in the optimum shift per switch.
@@
 \end{center}
+
 %TC:endignore
-
-\newpage
-
-\begin{center}
-    Cover Letter
-\end{center}
-
-Dear Editor,
-
-We are pleased to submit our manuscript entitled ”Phylogram for testing neutral evolution of a trait, chronogram to model selection” for consideration for publication in Evolution Letters.
-
-In this study, we revisit a puzzling question of phylogenetic comparative methods: does the time that have passed since the species diverged explain differences in traits?
-Or is it the genetic differences between species that explains best the differences?
-While the de facto standard in PCM is to use the time to explain the differences in traits, this reasoning is not always justified.
-Indeed, if the trait is evolving neutrally, we argue that the genetic differences between species instead of time should explain best the differences in traits.
-As a result, in this study, we show and argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.
-
-We believe that our manuscript will be of great interest to the readers of Evolution Letters as it offers a novel perspective on trait evolution and provides practical guidelines for researchers in the field.
-We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.
-
-With best regards,
-
-Thibault Latrille, Thibault Gaboriau and Nicolas Salamin
-
 \end{document}

### 2025-03-17 copy_artworks.sh

@@ -1,6 +1,8 @@
-EXT=pdf
-# Constant generation time
+EXT=png
+#/Users/tlatrille/Documents/venv/py312stats/bin/python3 /Users/tlatrille/Documents/ChronoPhylogram/scripts/plot_simulations_RevBayes.py --folder /Users/tlatrille/Documents/ChronoPhylogram/data_RevBayes/mammals_no --output /Users/tlatrille/Documents/ChronoPhylogram/results/mammals_no/inference_RevBayes.tsv
+#/Users/tlatrille/Documents/venv/py312stats/bin/python3 /Users/tlatrille/Documents/ChronoPhylogram/scripts/plot_simulations_RevBayes.py --folder /Users/tlatrille/Documents/ChronoPhylogram/data_RevBayes/mammals_slow --output /Users/tlatrille/Documents/ChronoPhylogram/results/mammals_slow/inference_RevBayes.tsv
 rm -rf ./manuscript/figures/simulation_*
+# Constant generation time
 cp ./results/mammals_no/inference_RevBayes.boxplot.is_nuc.${EXT} ./manuscript/figures/simulation_cstGT_phylogram_support.${EXT}
 cp ./results/mammals_no/inference_RevBayes.boxplot.is_OU.${EXT} ./manuscript/figures/simulation_cstGT_OU.${EXT}
 # Variable generation time
@@ -9,7 +11,7 @@ cp ./results/mammals_slow/inference_RevBayes.boxplot.is_OU.${EXT} ./manuscript/f
 cp ./results/mammals_slow/inference_RevBayes.boxplot.var_multiplier.${EXT} ./manuscript/figures/simulation_varGT_MultiBM.${EXT}
 cp ./results/mammals_slow/inference_RevBayes.boxplot.num_rate_changes.${EXT} ./manuscript/figures/simulation_varGT_MultiBM_NSwitch.${EXT}
 cp ./results/mammals_slow/inference_RevBayes.boxplot.num_theta_changes.${EXT} ./manuscript/figures/simulation_varGT_MultiOU_NSwitch.${EXT}
-cp ./results/mammals_slow/plot_RevBayesDistance_Phylo_neutral.${EXT} ./manuscript/figures/simulation_varGT_dist_NeutralPhylo.${EXT}
[…]

### 2025-03-17 run_all.sh

@@ -1,7 +1,13 @@
-for FILE in config/simulated_*.yaml;
+# Empirical data
+#snakemake --unlock
+#snakemake -j 10 -k --rerun-incomplete
+# Simulated data
+for FILE in config/simulated_mammals_te*.yaml;
 do
   echo "Running ${FILE}"
   cp -f "${FILE}" config/config.yaml
   snakemake --unlock
   snakemake -s workflow/Simulations.smk -j 10 -k --rerun-incomplete
 done
+# Copy figures
+./copy_artworks.sh

### 2025-03-17 scripts/plot_simulations_RevBayes.py

@@ -133,8 +133,8 @@ def main(folder, output):
                   "var_rates": ("Variance in the rates", "log", None, "v"),
                   "num_theta_changes": ("Number of optimum changes", "linear", 1.0, "m"),
                   "theta_multiplier": ("theta", "log", None, "n"),
-                  "is_BM": ("Probability of BM\n and not OU", "uniform", 0.5, "p_{BM}"),
-                  "is_OU": ("Probability of OU\n and not BM", "uniform", 0.5, "p_{OU}"),
+                  "is_BM": ("Support for BM over OU", "uniform", 0.5, "p_{BM}"),
+                  "is_OU": ("Support for OU over BM", "uniform", 0.5, "p_{OU}"),
                   "is_nuc": ("Support for a phylogram", "uniform", 0.5, "\\pi"),
                   "sigma": ("sigma", "log", None, "\\sigma"), "theta": ("theta", "linear", None, "\\theta"),
                   "alpha": ("alpha", "log", None, "\\alpha"), "t_half": ("t 1/2", "log", None, "T_{1/2}")}

### 2025-03-26 manuscript/main.tex
Updated manuscript with Lucy, Anna and Diego's comments

@@
 
-\renewcommand{\baselinestretch}{1.4}
+\renewcommand{\baselinestretch}{1.3}
 \renewcommand{\arraystretch}{1.2}
@@
 
+\begin{center}
+    \Huge Cover letter
+\end{center}
+{\Large
+Dear Editor,\\
+
+We are pleased to submit our manuscript entitled ”Phylogram for testing neutral evolution of a trait, chronogram to model selection” for consideration for publication in Evolution Letters.\\
+
+In this study, we revisit a puzzling question of phylogenetic comparative methods: does the time that have passed since the species diverged explain differences in traits?
+Or is it the genetic differences between species that explains best the differences?
+While the de facto standard in PCM is to use the time to explain the differences in traits, this reasoning is not always justified.
+Indeed, if the trait is evolving neutrally, we argue that the genetic differences between species instead of time should explain best the differences in traits.
+As a result, in this study, we show and argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.\\
+
+We believe that our manuscript will be of great interest to the readers of Evolution Letters as it offers a novel perspective on trait evolution and provides practical guidelines for researchers in the field.
+We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.\\
+
+With best regards,\\
+
+Thibault Latrille, Thibault Gaboriau \& Nicolas Salamin
+}
+\newpage
 \maketitle
@@
 On the other hand, neutral evolution is modelled as the trait changing randomly along the different branches of the species tree.
-But regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
+Regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
 Rationally, the time that has passed correlates with trait changes, even under a neutral regime, through the direct proxy of the number of generations that occurred.
-However, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases.
+However, we argue that since species generation time can also vary along the phylogenetic tree, chronograms introduces biases.
 Instead, for a phylogram, the branch lengths represent sequence divergence (i.e.
 the number of substitutions) which absorb the effect of changing generation time.
-Passed on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a random evolution has more support on a phylogram than on a chronogram.
+Based on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a random evolution has more support on a phylogram than on a chronogram.
 However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions.
@@
 When we look at different species, we can see that some traits are different between them.
-For example, the size of the brain or the expression level of a gene varies between species.
-The question we ask is: do the time that have passed since the species diverged explain the difference in traits?
-Or is it the genetic differences between species that explains best the differences?
-The standard in evolutionary biology is to use the time to explain the differences in traits.
-However, if the trait is evolving neutrally, the genetic differences between species instead of time should explain best the differences in traits.
-As a result, we argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.
+For example, the peak size of Darwin's finches varies between species.
+The question we ask is: does the species divergence time explain the differences in traits?
+Or is it the genetic differences between the species that best explains the differences?
+The standard in evolutionary biology is to use time to explain the differences in traits.
+However, if the trait is not under natural selection and it is instead changing due to accumulating mutations, the genetic differences between species instead of time should best explain the differences in traits.
+As a result, we argue that using simultaneously both the genetic differences and the species divergence time provides a better understanding of how traits evolve across species.
 
@@
 % The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
-By observing the changes in trait along a lineage, it is possible to speculate on the regime of selection acting on it.
-But even without selection, a trait might still changes due to random genetic drift.
-As such distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
-At a larger scale, from the observed variations of traits across species, such regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
+By observing the changes in a trait along a lineage, it is possible to speculate on the regime of selection acting on it.
+But even without selection, a trait might still change due to random genetic drift.
+As such, distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
+At a larger scale, from the observed variations of traits across species, regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
 For example, to model neutral evolution, the mean trait value is said to follow a Brownian motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
-Or in other words, for each  branch of the tree the value at descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
-In this framework, reconstructing the trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.
+In other words, for each branch of the tree, the value at the descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
+In this framework, reconstructing trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.
 
@@
 Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
-More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, when trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
-Methodologically, such alternative models of evolution raise issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
-Finally, the optimum can also be allowed to change continuously along the tree, which is interpreted as a trait tracking a moving optimum~\citep{hansen_translating_1996, latrille_detecting_2024}.
-One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
-This modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a continuously moving optimum.
+More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, where trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
+Methodologically, this alternative model of evolution raises issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
+Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a few abrupt changes in the environment along the phylogeny~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
+Finally, the optimum can also be allowed to change continuously along the tree, which results in the trait itself reflecting the movement of the optimum along the lineages due to constant adaptive evolution of the trait toward the optimum~\citep{hansen_translating_1996, hansen_three_2024}.
+One special case of a continuously changing optimum is again the BM, where changes in optimum are the consequences of a randomly changing environment~\citep{hansen_comparative_2008}, alternatively named fluctuating selection~\citep{holstad_evolvability_2024} or also diversifying selection due to the diversity of generated phenotypes at the clade level~\citep{latrille_detecting_2024}.
+Regardless of this discrepancy in terminology, this modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a continuously moving optimum.
 
-% This comes from the fact that we use chronogram, and not phylogram.
-% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
-One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
-Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
-Theoretically, for a neutrally evolving trait, trait changes depends directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the generation time is constant.
-But, since species generation time might vary along the phylogenetic tree, modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
-Instead, on a phylogram, the branch lengths can represent another quantity than time, which can be used as the backbone to model trait evolution.
-For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}.
-From a genomic perspective, phylogram can also be obtained in unit of nucleotide divergence, which is equivalent to measuring the number of nucleotide substitutions occurring along a branch.
-Such a phylogram in unit of nucleotide divergence, if obtained from neutrally evolving genomic loci, would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
-Conversely, for a trait evolving under selection at the species level, using a chronogram would be more accurate to model mean trait changes since shifts in the optimum fitness peak is extrinsic to the species, and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
-% Finally, from a statistical perspective, using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g.
-molecular clocks) to estimate the branch lengths of chronograms.
+% Deriving the neutral expectation from standing variation
+In order to disentangle neutral evolution from selection, another approach is to contrast the observed rate of evolution to the neutral expectation~\citep{lande_genetic_1980}.
+The neutral expectation can be obtained if the underlying genetic architecture of the trait is known and the trait encoded by many loci of additive effects~\citep{barton_infinitesimal_2017, sella_thinking_2019}.
+Theoretically, can also relate changes in mean trait value along a lineage to the trait variance at the population scale~\citep{turelli_heritable_1984, felsenstein_phylogenies_1988a}.
+Generalized to many species at the phylogenetic scale, contrasting between and within-species variations allow to disentangle neutral evolution from selection~\citep{latrille_detecting_2024}.
+And more generally, comparing the rate of evolution at different timescales allow to infer the regime of evolution~\citep{hansen_three_2024, holstad_evolvability_2024}.
+However, trait variation at different timescales is not always available, and the genetic architecture of the trait is often unknown.
+Instead, to disentangle neutral evolution from selection in the context of phylogenetic comparative methods, we hereby focus on a blind spot of the framework: the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
 
@@
 
+% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
+Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
+Theoretically, for a neutrally evolving trait, trait changes depend directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the average time between two consecutive generations, called generation time, is constant and the same for each species.
+Since generation time depends on the time for an individual to reach sexual maturity, it vary considerably between species (e.g.
+from a few months for coral reef pygmy gobies to 150 years for Greenland shark).
+As a result, variations of generation time between species means that modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
+
+% Phylogram to model neutral trait evolution
+Alternatively, on a phylogram, the branch lengths represent another quantity than time, which can be used as the backbone to model trait evolution and can potentially absorb changes in generation time.
+For example, it is known that using phylograms in units of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}, but in this case the phylogram is not expected to absorb changes in generation time.
+From a genomic perspective, phylograms can also be in units of nucleotide divergence, that is, depicting the number of nucleotide substitutions occurring along a branch.
+Because the number of neutral nucleotide substitutions accumulating along a branch is proportional to the number of generations~\citep{kimura_evolutionary_1968, kimura_neutral_1983}, nucleotide divergence would in this case absorb the effect of changing generation time (i.e.
+longer branches for lineages with shorter generation time).
+Such a phylogram, if obtained from neutrally evolving genomic loci would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
+Additionally, the rational to use nucleotide divergence to absorb changes in generation time is also valid for changes in mutation rate (per loci and per generation) under the assumption of a constant genetic architecture of the trait and that changes in mutation rate impact the whole genome~\citep{latrille_detecting_2024}.
+Empirically, such changes in mutation rate along a phylogenetic tree is also observed, although to a lesser extent than changes in generation time~\citep{bergeron_evolution_2023}.
+
+% Phylogram in the case of selection
+Under the alternative scenario of selection, shorter generation time and higher mutation rate can also allow for the species to track the changing optimum faster and not lag behind~\citep{lande_natural_1976, hansen_comparative_2008}.
+However, the timescale for this lag (hundred or thousands of years) is negligible compared to timescale on which the optimum changes at the level of clades (millions of years) in first approximation~\citep{hansen_three_2024}.
+Thus, at the phylogenetic scale, since shifts in the optimum fitness peak is extrinsic to the species and is dependent on the environment which varies with time, using a chronogram would be more accurate to model mean trait changes.
+As a result, for a trait under selection, mean trait changes should be better explained by the time rather than the number of substitutions that occurred along a branch.
+
 % Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
-In this study, we test if using a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram can allow to assess the regime of evolution of a trait (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
-More precisely, we seek to test if change of a neutral trait is better predicted a phylogram.
-Conversely, we also seek to test if changes of a trait under a moving optimum is better predicted by a chronogram.
-As such, by using genomic information we seek to provide additional approaches to model trait evolution, particularly to disentangle selection and neutral evolution.
+In this study, we test if using a phylogram (in units of nucleotide divergence for neutral loci) instead of a chronogram can allow to assess the regime of evolution of a trait (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
+More precisely, we seek to test if the changes of a neutral trait are better predicted by a phylogram.
+Conversely, we also seek to test if changes of a trait under a moving optimum are better predicted by a chronogram.
+Using genomic information we seek to provide additional approaches to model trait evolution, particularly to disentangle selection and neutral evolution.
 
@@
 Third, we fitted a single rate Brownian motion (BM) to the simulated data, and we compared the fit of the BM on a phylogram versus a chronogram.
-Forth, based in the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated and empirical datasets.
+Fourth, based in the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated and empirical datasets.
 Finally, we fitted several models of trait evolution to the simulated and empirical datasets: a multi-rate BM, an Ornstein-Uhlenbeck (OU) process, and a relaxed OU with multiple optima.
@@
 In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
-Parents were selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
-Flattening the fitness landscape result in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
-Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
-Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch, if a switch occurs the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).
+Parents were selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
+Flattening the fitness landscape resulted in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
+Alternatively, for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
+Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch; if a switch occurs, the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).
 
 At each node of the tree, the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached (Fig.~\ref{fig:results-brownian}C).
-As a control, we performed simulations with constant mutation rate, generation times and effective population sizes ($\Ne$).
-Alternatively, we performed simulations with fluctuating mutation rate, generation times and $\Ne$, were we used a BM to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).
+As a control, we performed simulations with constant mutation rates, generation times and effective population sizes ($\Ne$).
+Alternatively, we performed simulations with fluctuating mutation rates, generation times and $\Ne$, where we used a BM to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).
 
-The phylogram is obtained from the same simulation setting on a set of independent 30.000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
+The phylogram is obtained from the same simulation setting on a set of independent 30000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
 The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
@@
 The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
-More specifically, the phylogram in unit of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
+More specifically, the phylogram in units of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
 The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
@@
 On the simulated dataset, we fitted a single rate BM using either a phylogram or a chronogram as the underlying tree.
-We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian motion (BM) on continuous traits evolution evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
+We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian motion (BM) on continuous traits evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
 The data consist of the mean trait value for each extant species, and the tree topology is fixed.
 Along each branch, the value of the trait is drawn from a normal distribution with mean equal to the parent node value and variance equal to the branch length~\citep{felsenstein_phylogenies_1985}.
-We reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
+Formally, the BM process is described by the equation:
+% Ornstein-Uhlenbeck model equations
+\begin{align}
+    \der \Trait_t & = \sigma \Multiply \der W_t,  \label{eq:bm-process}
+\end{align}
+where $\Trait_t$ is the trait value at time $t$, $\sigma$ is the rate of the BM, and $\der W_t$ is a Wiener process (i.e. a standard Brownian motion).
+We used a log-uniform prior for $\sigma$.
+Using Markov chain Monte Carlo (MCMC), we reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
 Both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
 Importantly, the data and the tree topology are the same in both analyses, only the branch lengths are different.
-We expect for trait under neutral evolution more accurate trait reconstruction on a phylogram, and conversely for trait under selection (moving optimum) more accurate trait reconstruction on a chronogram.
+For traits simulated under neutral evolution, we expect more accurate trait reconstruction on a phylogram, and conversely, for traits under selection (moving optimum), more accurate trait reconstruction on a chronogram.
 
@@
 To test the support of a phylogram over a chronogram for the simulated and empirical datasets, we implemented a model based on a single rate BM.
-Such model was implemented in \textit{RevBayes} and contains a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
+The model was implemented in \textit{RevBayes} and contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
 Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
 If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
-The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
+The tree topology is fixed, meaning both trees have the same branching structure, but because branch lengths are different and not necessarily on the same scale, both trees are re-scaled by dividing branch lengths by the total tree length.
 As in the previous section, a BM is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
@@
 Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
-The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favoured over the chronogram.
-We expect $\PhyloSupport = 1$ for trait under neutral evolution, and $\PhyloSupport = 0$ for trait under selection (moving optimum).
+The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability that the phylogram is favoured over the chronogram.
+We expect $\PhyloSupport = 1$ for a trait under neutral evolution, and $\PhyloSupport = 0$ for a trait under selection (moving optimum).
 
-\subsection*{Alternatives to single rate BM}
-Additionally to a single rate BM, we fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
+\subsection*{Alternative models of evolution}
+In addition to a single rate BM, we  also fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
 The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
+For each branch, we draw a rate-multiplier either equal to $1$ (no rate shift), or drawn from a log-normal distribution with a median of 1, and a standard deviation such that rate shifts range over about one order of magnitude.
+We set to our prior expected number of rate shifts to $1$ across the whole tree.
 The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1000 gen., running of 50000 gen., 2 chains).
@@
 % Ornstein-Uhlenbeck model equations
+The OU process is described by the equation:
 \begin{align}
-    \der \Trait_t & = -\alpha \left( \Trait_t - \theta \right) \der t + \sigma \der W_t \label{eq:ou-process}
+    \der \Trait_t & = -\alpha \left( \Trait_t - \theta \right) \der t + \sigma \Multiply \der W_t, \label{eq:ou-process}
 \end{align}
-In the OU process, the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, and $\theta$ is the optimum value.
+where the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, $\theta$ is the optimum value and the other parameters are defined as in eq~\ref{eq:bm-process}.
 The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the BM, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
 % alpha ~ dnReversibleJumpMixture(0.0, dnExponential( abs(root_age / 2.0 / ln(2.0)) ), 0.5)
-Mathematically, $\alpha$ is drawn from a reversible-jump mixture distribution: a value of $0$ or drawn from an exponential distribution.
-Thus, when $\alpha = 0$ the OU process is equivalent to a BM.
+The prior for $\theta$ is uniform between the minimum and maximum trait value for extant species.
+The prior for $\alpha$ is a reversible-jump mixture distribution: a value of $0$ or drawn from an exponential distribution with mean equal to half the root age divided by $\ln (2)$, meaning that we expect a phylogenetic half life of half the tree age.
+Thus, when $\alpha = 0$ the OU process (eq.~\ref{eq:ou-process}) is equivalent to a BM (eq.~\ref{eq:bm-process}).
 The support for the OU model is estimated as the posterior mean of the switch variable ($p_{\text{OU}}$), or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).
@@
 The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1000 gen., running of 5000 gen., 2 chains).
+For each branch, we draw a shift in optimum either equal to $0$ (no shift), or drawn from a uniform distribution centred on $0$ with breadth spanning the whole range of observed trait value.
+We set to our prior expected number of optimum shifts to $1$ across the whole tree.
 
@@
     \centering
-    \includegraphics[width=\textwidth, page=1] {figures/artwork_simulations}\\
+    \includegraphics[width=\textwidth, page=1] {figures/artwork_pipeline}\\
     \vspace{10pt}
     \begin{minipage}{0.49\linewidth}
-        \flushleft {\small  \textbf{D: simulations with constant generation time}}
+        \flushleft {\small \textbf{C.\ Simulations with constant generation time}}
         \includegraphics[width=\linewidth, page=1]{figures/simulation_cstGT_phylogram_support}
@@
     \begin{minipage}{0.49\linewidth}
-        \flushleft {\small  \textbf{E: simulations with fluctuating generation time}}
+        \flushleft {\small \textbf{D.\ Simulations with fluctuating generation time}}
         \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_phylogram_support}
@@
     \caption{
-        Panel~A: Wright-Fisher simulations with mutation, selection and drift.
-        For a given individual, its genotypic value is encoded by a finite number of independent loci (meaning no linkage) contributing additively.
-        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
-        Panel~B: Neutral evolution is modelled as a constant fitness regardless of phenotypic value.
-        Selection is modelled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree.
-        Panel~C: Example simulation of a trait evolving along a phylogenetic tree.
-        Panels D \& E: Violinplot of posterior probability of the phylogram being favoured over the chronogram ($\PhyloSupport$) across 100 replicate simulations for the different simulated regime of evolution.
-        Horizontal lines inside violins are the estimated $\PhyloSupport$ of each replicate simulation.
-        Regime of evolution: neutral regime (yellow), under a moving optimum (blue), and under multiple optima (red).
-        Results of simulations with constant generation time (Panel~D) and with fluctuating generation time, mutation rate and effective population size (Panel~E).
+        \textbf{Testing the support for a phylogram over a chronogram}
+        Panel~A:
+        Test for the support of a phylogram over a chronogram using a Brownian motion (BM) to model trait changes along a phylogeny.
+        The model contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time ($\PhyloSupport=0$, chronogram in purple) to units of nucleotide divergence for neutral sites ($\PhyloSupport=1$, phylogram in green).
+        The posterior mean estimate of $\PhyloSupport$ indicates whether a BM supports better a phylogram or a chronogram ($0 \leq \PhyloSupport \leq 1$), given the mean trait value for each extant species and both trees (chronogram \& phylogram) with the same topology.
+        Only if the phylogram has more support than chronogram ($\PhyloSupport$ close to $1$) there is some support that the trait might evolve under drift.
+        Panel~B: Wright-Fisher simulations of trait evolution along a phylogeny.
+        Neutral evolution is modelled as a constant fitness regardless of the phenotype of individuals (yellow).
+        Selection is modelled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree: as a moving optimum (blue) or as multiple optima (red).
+        At each node of the tree we have a speciation event and the two descendant species evolve independently.
+        The output of the simulation is the mean trait value for each extant species, but we also save the phylogram and the chronogram as obtained from independent markers.
+        We then compute the posterior mean estimate of $\PhyloSupport$ from the output of the simulation.
+        Panels C \& D: Violin plot of the posterior mean of $\PhyloSupport$ across 100 replicates for the different simulated regimes of evolution with mean $\PhyloSupport$ values at the center ($\MeanPhyloSupport$).
+        Horizontal lines inside the violins are the estimated $\PhyloSupport$ of each replicate simulations.
+        Results of simulations with constant generation time (Panel~C) and with fluctuating generation time, mutation rate and effective population size (Panel~D).
     }
@@
 We first tested the hypothesis that a phylogram is better suited to model neutral evolution than a chronogram.
-We simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum and 3) under multiple optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
-We tested for the support of phylogram over a chronogram under the assumption that changes follow a Brownian motion (BM) process, we developed as statistic denoted as $\PhyloSupport$, which range from 0 (full support for the chronogram) to 1.0 (full support for the phylogram).
-For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
+We simulated traits evolving on a phylogenetic tree under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum, and 3) under multiple optima (see Methods, Fig.~\ref{fig:results-brownian}A-C).
+We first assessed the accuracy of ancestral trait reconstruction on a phylogram versus a chronogram, under the assumption that changes follow a Brownian motion (BM).
+For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
+In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$) than on a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).
+
+We then tested for the BM support of phylogram over a chronogram, and we developed as statistic denoted as $\PhyloSupport$, which ranges from 0 (full support for the chronogram) to 1 (full support for the phylogram).
+For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch lengths are equivalent in both trees.
 Next, we simulated changing generation time, mutation rate (per generation) and effective population size ($\Ne$) along the phylogenetic tree, mimicking a mammalian range of changes.
-In this case, for a neutral trait, the BM was better fitted on a phylogram than on a chronogram (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
-Conversely, for a trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
+In this case, for a simulated neutral trait, the BM was better fitted on a phylogram than on a chronogram, with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold (Fig.~\ref{fig:results-brownian}E, yellow).
+Conversely, for a simulated trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}E, blue) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}E, red).
 
-On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
-When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
-For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
-On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
-We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).
+On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$, when sex is not taken into account.
+When splitting the dataset into males and females, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
+For brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
+On the \textit{COMBINE} dataset, which does not distinguish between males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
+We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), a method that allow for faster computation and scale better for larger trees, and found similar results (Table~\ref{table:results-empirical}).
+
+\subsection*{Fitting alternative models of evolution}
+% Mis-classification of trait evolution on a phylogram.
+Besides testing the fit of a BM with a single rate, we also fitted a multi-rate BM (see Methods).
+When fitting a multi-rate BM, the estimated variance of rate parameters ($v$) was lower on a phylogram than on a chronogram for a trait evolving neutrally (Fig.~\ref{fig:results-alternative}A, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.8\times 10^{-14}$), and the number of rate shifts ($n$) was reflecting the prior on the phylogram while not on the chronogram, showing a bias (Fig.~\ref{fig:results-alternative}B, yellow violins).
+Conversely, for a trait under a moving optimum, $v$ was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}A, blue violins, Wilcoxon paired test with  $p_{\text{value}}=8.9\times 10^{-15}$), and $n$ was reflecting the prior on the chronogram while not on the phylogram, showing a bias (Fig.~\ref{fig:results-alternative}B, blue violins).
+Altogether, fitting a BM on a phylogram is more accurate for a trait evolving neutrally, but results in less accurate estimates for a trait under selection.
+
+% Mainly testing BM versus OU
+Moreover, we tested alternatives to the BM with Ornstein-Uhlenbeck (OU) models: with a single optimum and with multiple optima (see Methods).
+First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\text{value}}=3.9\times 10^{-18}$), an expected result since the BM with a switch supported the chronogram.
+For a trait evolving neutrally, $p_{\text{OU}}$ was still higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.0\times 10^{-8}$), an unexpected result since the BM with a switch supported the phylogram, thus showing a high rate of mis-classification on a phylogram.
+When fitting a multi-OU process on a trait evolving neutrally, the number of optimum shifts ($m$) was reflecting the prior on the phylogram, but not on the chronogram (Fig.~\ref{fig:results-alternative}B, yellow violins).
+Conversely, for a trait under a moving optimum, ($m$) was reflecting the prior on the chronogram but not on the phylogram (Fig.~\ref{fig:results-alternative}A, blue violins).
+In summary, fitting an OU process on a phylogram results in biases not only for a trait under selection but also for a trait under neutral evolution.
 
@@
         \textbf{Mis-classification of trait evolution on a phylogram.}
-        Violinplot of posterior parameter estimates for different models of trait evolution and different simulated regime of selection.
-        Horizontal lines inside the violin show the result of each replicate simulation.
+        Violin plot of posterior parameter estimates for different models of trait evolution and different simulated regimes of selection.
+        Horizontal lines inside the violin show the results of each replicate simulation.
         Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow) and under a moving optimum (blue).
@@
 
-\subsection*{Trait evolution on a phylogram and on a chronogram}
-% Mis-classification of trait evolution on a phylogram.
-Beside support for the underlying tree, we also assessed the accuracy of ancestral trait reconstruction on a phylogram versus a chronogram.
-For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
-In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$) than on a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).
-Additionally, instead of testing the fit of a BM with a single rate, we also fitted a multi-rate BM process (see Methods).
-When fitting a multi-rate BM process, the estimated variance of rate parameters ($v$) was lower on a phylogram than on a chronogram for a trait evolving neutrally (Fig.~\ref{fig:results-alternative}A, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.8\times 10^{-14}$), and the number of rate shifts ($n$) was reflecting the prior on the phylogram while not on the chronogram, showing a bias (Fig.~\ref{fig:results-alternative}B, yellow violins).
-Conversely, for a trait under a moving optimum, $v$ was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}A, blue violins, Wilcoxon paired test with  $p_{\text{value}}=8.9\times 10^{-15}$), and $n$ was reflecting the prior on the chronogram while not on the phylogram, showing a bias (Fig.~\ref{fig:results-alternative}B, blue violins).
-Altogether, fitting a BM on a phylogram is more accurate for a trait evolving neutrally, but results in less accurate estimates for a trait under selection.
-
-% Mainly testing BM versus OU
-Moreover, we tested alternative to the BM process with Ornstein-Uhlenbeck (OU) models: with a single optimum and with multiple optima (see Methods).
-First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\text{value}}=3.9\times 10^{-18}$).
-Worst still, even for a trait evolving neutrally, $p_{\text{OU}}$ was still higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.0\times 10^{-8}$).
-When fitting a multi-OU process on a trait evolving neutrally, the number of optimum shifts ($m$) was reflecting the prior on the phylogram while not on the chronogram (Fig.~\ref{fig:results-alternative}B, yellow violins).
-Conversely, for a trait under a moving optimum, ($m$) was reflecting the prior on the chronogram while not on the phylogram (Fig.~\ref{fig:results-alternative}A, blue violins).
-Altogether, fitting an OU process on a phylogram results in biases, for a trait under selection but even for a trait under neutral evolution.
-
 \section*{Discussion}
@@
 In phylogenetic comparative methods, the evolution of continuous traits is typically modelled as a stochastic process running along the branches of a phylogenetic tree.
-The null model of neutral trait evolution, genetic drift, is though as a Brownian motion (BM) running on the tree~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
-As a result, traits on which the BM is favoured over alternative models are sometimes interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
+The null model of neutral trait evolution, genetic drift, is thought of as a Brownian motion (BM) running on the tree~\citep{lynch_phenotypic_1986, felsenstein_phylogenies_1988a}, and deviations from this model are interpreted as selection acting on the trait~\citep{butler_phylogenetic_2004}.
+As a result, a trait on which the BM is favoured over alternative models are sometimes interpreted as a trait evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019}.
 However, the BM is not only the null model of neutral evolution, but can also model selection, with a trait tracking a moving optimum, where the optimum itself is following a BM~\citep{hansen_translating_1996, latrille_detecting_2024}.
-Such identifiability is problematic, but we here argue that the underlying backbone tree to model trait evolution can help to disentangle neutral evolution from selection.
-The standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are measured in time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}, while phylograms have been generally overlooked, where the branch lengths are measured in number of substitutions instead.
-Such overlook is rational and justified, since for simulated trait under selection we assessed that the BM has better fit on a chronogram than on a phylogram.
-However, we here argue that to model a trait evolving neutrally, the backbone tree should not be a chronogram but instead a phylogram.
-Such phylogram would better represent the number of generations that occurred, and in turn should explain best the differences in traits between species~\citep{latrille_detecting_2024}
+Distinguishing these scenarios is not trivial, but we here showed that the underlying backbone tree to model trait evolution can help disentangle neutral evolution from selection.
+The standard in phylogenetic comparative methods is a chronogram, where the branch lengths are measured in time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}, while phylograms, where the branch lengths are measured in number of substitutions instead, have been generally overlooked.
+For a simulated trait under selection we assessed that the BM has indeed a better fit on a chronogram than on a phylogram.
+However, we argue that to model a trait evolving neutrally, the backbone tree should not be a chronogram but instead a phylogram.
+Phylograms would better represent the number of generations that occurred, and in turn should explain best the differences in traits between species~\citep{latrille_detecting_2024}
 In practice, we show that for simulations of a trait evolving neutrally, the fit of a BM supports a phylogram rather than a chronogram.
-Additionally, for the neutral trait, both ancestral state reconstruction and estimation for the number of rate changes for a multi-rate BM was also more accurate on a phylogram.
-Altogether, the combined use of a phylogram and a chronogram can help support the hypothesis of neutral evolution, or instead rule it out.
+Additionally, for a neutral trait, both ancestral state reconstruction and estimation for the number of rate changes for a multi-rate BM was more accurate on a phylogram.
+Altogether, the combined use of a phylogram and a chronogram can help support the hypothesis of neutral evolution, or to rule it out.
 % Thus, we argue that claims for a trait evolving under drift should be tested more thoroughly by comparing the fit of the BM under both a phylogram and a chronogram as the underlying tree.
@@
 % When does it apply and why is it important
-Applied to different empirical dataset of body and brain masses in mammals, we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these trait~\citep{latrille_detecting_2024}.
-The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree.
+We applied this method to different empirical datasets of body and brain masses in mammals, and we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these traits~\citep{latrille_detecting_2024}.
+The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree (e.g. from 6 months for mice to 50 years for bowhead whales).
 But more generally, our method is also relevant when changes in mutation rate (per generation) and effective population size ($\Ne$) are expected.
-First, normalization by nucleotide divergence also accounts theoretically for variations in effective population size ($\Ne$) since the substitution rate of neutral mutations is equal to the mutation rate, such that $\Ne$ has no effect on the rate of neutral evolution~\citep{kimura_evolutionary_1968, ohta_population_1972}.
-Moreover, if population structure impacts the probability of fixation for neutral mutations (that would also impact a neutral trait), the phylogram would absorb automatically these changes.
-This argument is not only verbal since we tested for it in our simulation by modelling changes in both $\Ne$ and generation time, and the phylogram was favoured over the chronogram for a trait evolving neutrally.
-Second, aside from absorbing of changes in generation time and $\Ne$, the use of phylograms also absorbs changes in mutation rate (per generation) under certain assumptions.
-As is the case in our simulations, we assumed that the genetic architecture of the trait is constant and the mutation rate is homogeneous across the genome.
-However, if the mutation rate would only changes for a subset of the genome that is encoding the trait, the phylogram would not absorb these changes in mutation rate.
-Finally, because chronogram and usually derived from chronogram by calibrating the tree with molecular clocks, using phylogram has the additional advantage of removing model assumptions required to calibrate node ages~\citep{litsios_effects_2012}.
+First, normalizing by nucleotide divergence also accounts theoretically for variations in effective population size ($\Ne$) since the substitution rate of neutral mutations is equal to the mutation rate, such that $\Ne$ has no effect on the rate of neutral evolution~\citep{kimura_evolutionary_1968, ohta_population_1972}.
+Moreover, if population structure were to impact the probability of fixation for neutral mutations (which would also impact a neutral trait), the phylogram would automatically absorb these changes.
+This argument was partially tested in our simulation by modelling changes in both $\Ne$ and generation time (although not modelling population structure \textit{per se}), and the phylogram was favoured over the chronogram for a trait evolving neutrally.
+Second, under certain conditions, the use of phylograms can also absorbs changes in mutation rate (per loci per generation), which in mammals can vary up to 10 fold~\citep{bergeron_evolution_2023}.
+As in the case of our simulations, we assumed that the genetic architecture of the trait is constant and the mutation rate is homogeneous across the genome.
+However, if the mutation rate would only change for a subset of the genome that is encoding the trait, the phylogram would not absorb these changes in mutation rate.
+Finally, because chronogram are usually derived from phylogram by calibrating the tree with molecular clocks, using phylograms directly has the additional advantage of removing model assumptions required to calibrate node ages~\citep{litsios_effects_2012}.
 
 % When does it fall appart, our second result
-These examples show that while phylograms are useful to neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
-In our simulations, we found that generally using a phylogram tend to favour the OU process over the Brownian motion (BM) model.
-This suggests that trees in unit of nucleotide divergence may not always provide a reliable framework for modelling trait evolution, especially when selection is involved.
-As such, we recommend continuing to use chronograms for modelling selection acting on a trait, as they generally provide more accurate and reliable results.
-Moreover, for a trait is under selection for which the pace of optimum changes is also tracking the changes in generation time, the support for a phylogram over a chronogram would be misleading~\citep{litsios_effects_2012}.
+These examples show that while phylograms are useful to model neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
+In our simulations, we found that generally using a phylogram tends to favour the OU process over the BM model.
+This suggests that trees in units of nucleotide divergence may not always provide a reliable framework for modelling trait evolution, especially when selection is involved.
+We recommend continuing to use chronograms for modelling selection acting on a trait, as they generally provide more accurate and reliable results.
+Moreover, for a trait under selection for which the pace of optimum changes is also tracking the changes in generation time, the support for a phylogram over a chronogram would be misleading~\citep{litsios_effects_2012}.
 Therefore, we argue that while phylograms can offer valuable insights~\citep{wilson_chronogram_2022}, they should be used with caution and in conjunction with chronograms to provide a more comprehensive understanding of trait evolution.
@@
 Caution is indeed warranted, as there are several limitations to our approach.
-First, we simulated only a subset of different regimes of selection, but more complex scenarios could be included for example with with primary and secondary moving optimums~\citep{hansen_three_2024}.
+First, we simulated only a subset of different regimes of selection, but more complex scenarios could be included, for example with primary and secondary moving optimums~\citep{hansen_three_2024}.
 Additionally, the simulated genetic architecture of a single trait is assumed to be constant across the phylogeny.
-Such constant architecture result in saturation of the phenotype on long time scales, with a phenomenon similar to nucleotide saturation~\citep{latrille_detecting_2024}.
-Such saturation might be one of the reason for a bias in favour of the Ornstein-Uhlenbeck (OU) process over the Brownian motion (BM) model that we observed in our simulations.
+Constant architecture results in the saturation of the phenotype on long time scales, with a phenomenon similar to nucleotide saturation~\citep{latrille_detecting_2024}.
+This saturation effect might be one of the reasons for a bias in favour of the Ornstein-Uhlenbeck (OU) process over the BM model that we observed in our simulations.
 A more realistic genetic architecture should include pleiotropy, epistasis and linkage disequilibrium, while here we only considered non-linked loci contributing additively to a single trait.
-Second, while the tested alternative models of evolution (multi-rates BM, OU, multi-optimum OU) are standard, they do not encompass the full breadth of available methods, which could be expanded~\citep{pennell_geiger_2014, hohna_revbayes_2016}.
-Finally, the dataset of body and brain masses in mammals may not be representative of all traits or species and even though our analysis showcase the utility of phylograms, more empirical studies are needed to replicate the findings.
+Second, while the tested alternative models of evolution (multi-rates BM, OU, multi-optimum OU) are standard, they do not encompass the full breadth of available methods to model trait evolution~\citep{pennell_geiger_2014, hohna_revbayes_2016}.
+Finally, the mammalian dataset (brain and body) may not be representative of all traits or species and even though our analysis showcase the utility of phylograms, more empirical studies are needed to replicate the findings.
 
 % Gene expression level and empirical studies.
-One striking example of traits that could (and should) be empirically studied with a phylogram is gene expression level, both for protein concentration and mRNA expression.
-Indeed, the regime of selection acting on gene expression level is the focus of intense debates~\citep{signor_evolution_2018, price_detecting_2022, bertram_cagee_2023, dimayacyac_evaluating_2023}.
-And typically, dataset on which the BM is favoured over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
+One striking example of a trait that could (and should) be empirically studied with a phylogram is gene expression level, both for protein concentration and mRNA expression.
+Indeed, the regime of selection acting on gene expression level is the focus of intense debate~\citep{signor_evolution_2018, price_detecting_2022, bertram_cagee_2023, dimayacyac_evaluating_2023}.
+Typically, datasets on which the BM is favoured over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
 Phenotypic distance is also plotted as a function of time while claiming that a trait is evolving under drift~\citep{jiang_decoupling_2023}.
@@
 % Conclusion
-Altogether, our study support the use a chronogram when testing for selection acting on a trait.
+Altogether, our study supports the use of a chronogram when testing for selection acting on a trait.
 The pipeline of hypothesis testing and model comparison is currently well established and should be followed.
 However, when the analysis concludes that the BM is favoured, then the support of the phylogram over the chronogram should be tested additionally.
-We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis, which indicate whether a BM support better a phylogram or a chronogram ($0 \leq \PhyloSupport \leq 1$), given data at the tip of the tree and both trees with the same topology.
-Only then can we claim that a trait is evolving under drift if the phylogram has more support than chronogram ($\PhyloSupport$ close to $1$), but caution is still warranted.
+We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis, which indicates whether a BM supports better a phylogram or a chronogram ($0 \leq \PhyloSupport \leq 1$), given the data at the tip of the tree and both trees with the same topology.
+Only if the phylogram has more support than chronogram ($\PhyloSupport$ close to $1$) there is some support that the trait might evolve under drift, but caution is still warranted.
 
@@
 \label{sec:acknowledgment}
-We gratefully acknowledge the help of Julien Joseph for his advice and reviews concerning this manuscript.
+We gratefully acknowledge the help of Lucy M. Fitzgerald, Diego A. Hartasánchez, Anna Marcionetti and Julien Joseph for their advice and reviews concerning this manuscript.
 

### 2025-08-14 manuscript/main-first-submission.tex
Added manuscript tex file and prepare for revision

--- 2025-03-26 manuscript/main.tex
@@
 
-% \title{Phylograms disentangle drift and selection acting on a continuous trait.}
-% \title{Drift and selection acting on a continuous trait can be disentangled by a simultaneous use of a chronogram and a phylogram}
-% \title{Phylogram instead of chronogram when claiming that a trait is evolving under drift.}
-\title{Phylogram for testing neutral evolution of a trait, chronogram to model selection.}
-\rhead{\scshape Chronogram or phylogram for trait evolution.}
+\title{Phylogram instead of chronogram when assessing the neutral evolution of a trait}
+\rhead{\scshape Phylogram over chronogram for neutral evolution.}
 
@@
 \normalsize
-$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
+$^{1}$Department of Computational Biology, University of Lausanne, Quartier Sorge, 1015 Lausanne, Switzerland\\
 \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
@@
 %TC:ignore
-
-\begin{center}
-    \Huge Cover letter
-\end{center}
-{\Large
-Dear Editor,\\
-
-We are pleased to submit our manuscript entitled ”Phylogram for testing neutral evolution of a trait, chronogram to model selection” for consideration for publication in Evolution Letters.\\
-
-In this study, we revisit a puzzling question of phylogenetic comparative methods: does the time that have passed since the species diverged explain differences in traits?
-Or is it the genetic differences between species that explains best the differences?
-While the de facto standard in PCM is to use the time to explain the differences in traits, this reasoning is not always justified.
-Indeed, if the trait is evolving neutrally, we argue that the genetic differences between species instead of time should explain best the differences in traits.
-As a result, in this study, we show and argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.\\
-
-We believe that our manuscript will be of great interest to the readers of Evolution Letters as it offers a novel perspective on trait evolution and provides practical guidelines for researchers in the field.
-We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.\\
-
-With best regards,\\
-
-Thibault Latrille, Thibault Gaboriau \& Nicolas Salamin
-}
-\newpage
 \maketitle
@@
 To disentangle these processes, different scenarios of evolution are modelled and compared.
-For traits under selection, the species is seen as tracking a trait optimum, and such optimum can change along the species tree.
-On the other hand, neutral evolution is modelled as the trait changing randomly along the different branches of the species tree.
-Regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
-Rationally, the time that has passed correlates with trait changes, even under a neutral regime, through the direct proxy of the number of generations that occurred.
-However, we argue that since species generation time can also vary along the phylogenetic tree, chronograms introduces biases.
-Instead, for a phylogram, the branch lengths represent sequence divergence (i.e.
-the number of substitutions) which absorb the effect of changing generation time.
-Based on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a random evolution has more support on a phylogram than on a chronogram.
-However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions.
+For quantitative traits under selection, the species is usually considered to track a trait optimum, and such an optimum can change along the branches of the species tree.
+On the other hand, neutral evolution is modelled with a trait changing randomly around the ancestral trait value along the different branches of the species tree.
+Regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently of the trait data analysed.
+Branch lengths are usually assumed to be in units proportional to time, and the tree is represented by a chronogram.
+The rationale is that time correlates with trait changes because of its direct connection with the number of generations that occurred.
+However, since the generation time of species can also vary along the phylogenetic tree, we argue that their use introduces biases.
+In contrast, if the phylogenetic tree is represented by a phylogram with branch lengths in units of sequence divergence, this will account for the effect of changing generation time.
+In this study, we show using simulations that, for a trait evolving neutrally, the fit of a random evolution model has more support on a phylogram than on a chronogram.
+However, comparing models and testing different scenarios of selection using a phylogram leads to incorrect predictions.
 Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift.
-Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a trait.
+Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a quantitative trait.
 \end{abstract}
@@
 When we look at different species, we can see that some traits are different between them.
-For example, the peak size of Darwin's finches varies between species.
-The question we ask is: does the species divergence time explain the differences in traits?
-Or is it the genetic differences between the species that best explains the differences?
-The standard in evolutionary biology is to use time to explain the differences in traits.
-However, if the trait is not under natural selection and it is instead changing due to accumulating mutations, the genetic differences between species instead of time should best explain the differences in traits.
-As a result, we argue that using simultaneously both the genetic differences and the species divergence time provides a better understanding of how traits evolve across species.
+For example, the brain size of different primate and human lineages varies.
+The question we ask is: does the time that passed explain the differences in their traits?
+Or are genetic differences between species better at explaining the differences?
+The standard in evolutionary biology is to use time to explain differences in traits.
+However, if the trait is not under natural selection and instead is changing due to accumulating mutations, the genetic differences between species instead of time should best explain the differences in traits.
+As a result, we argue that using both genetic differences and species divergence time simultaneously provides a better understanding of how traits evolve across species.
 
@@
 % The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
-By observing the changes in a trait along a lineage, it is possible to speculate on the regime of selection acting on it.
-But even without selection, a trait might still change due to random genetic drift.
-As such, distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
+By measuring changes in a phenotypic trait along a lineage, it is possible to infer the type of selection acting on it or whether this change is only due to random genetic drift.
 At a larger scale, from the observed variations of traits across species, regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
@@
 % Alternative models of trait evolution and their pitfalls.
-Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
+Alternatively to a simple BM, adding a trend parameter in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
 More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, where trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
 Methodologically, this alternative model of evolution raises issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
-Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a few abrupt changes in the environment along the phylogeny~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
-Finally, the optimum can also be allowed to change continuously along the tree, which results in the trait itself reflecting the movement of the optimum along the lineages due to constant adaptive evolution of the trait toward the optimum~\citep{hansen_translating_1996, hansen_three_2024}.
-One special case of a continuously changing optimum is again the BM, where changes in optimum are the consequences of a randomly changing environment~\citep{hansen_comparative_2008}, alternatively named fluctuating selection~\citep{holstad_evolvability_2024} or also diversifying selection due to the diversity of generated phenotypes at the clade level~\citep{latrille_detecting_2024}.
-Regardless of this discrepancy in terminology, this modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a continuously moving optimum.
+Adding biological realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a few abrupt changes in the environment along the phylogeny~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
+Finally, the optimum can also be allowed to change continuously along the tree, which results in the trait itself reflecting the movement of the optimum along the lineages due to the constant adaptive evolution of the trait toward the optimum~\citep{hansen_translating_1996, hansen_three_2024}.
+One special case of a continuously changing optimum is again the BM, where changes in the optimum are a consequence of the randomly changing environment~\citep{hansen_comparative_2008}, named fluctuating selection~\citep{holstad_evolvability_2024} or also diversifying selection due to the diversity of generated phenotypes at the clade level~\citep{latrille_detecting_2024}.
+This modelling raises an issue in disentangling selection from neutral evolution: the best fit of BM can not be interpreted as the trait evolving neutrally, it can also be interpreted as a continuously moving optimum.
 
@@
 The neutral expectation can be obtained if the underlying genetic architecture of the trait is known and the trait encoded by many loci of additive effects~\citep{barton_infinitesimal_2017, sella_thinking_2019}.
-Theoretically, can also relate changes in mean trait value along a lineage to the trait variance at the population scale~\citep{turelli_heritable_1984, felsenstein_phylogenies_1988a}.
-Generalized to many species at the phylogenetic scale, contrasting between and within-species variations allow to disentangle neutral evolution from selection~\citep{latrille_detecting_2024}.
-And more generally, comparing the rate of evolution at different timescales allow to infer the regime of evolution~\citep{hansen_three_2024, holstad_evolvability_2024}.
+Additionally, we can also derive the expected changes in mean trait value along a lineage given the trait variance at the population scale~\citep{turelli_heritable_1984, felsenstein_phylogenies_1988a}.
+When generalized to many species at the phylogenetic scale, contrasting between and within-species variations allow us to disentangle neutral evolution from selection~\citep{latrille_detecting_2024}.
+More generally, comparing the rate of evolution at different timescales allows inferring the regime of evolution~\citep{hansen_three_2024, holstad_evolvability_2024}.
 However, trait variation at different timescales is not always available, and the genetic architecture of the trait is often unknown.
-Instead, to disentangle neutral evolution from selection in the context of phylogenetic comparative methods, we hereby focus on a blind spot of the framework: the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
+Instead, to disentangle neutral evolution from selection in the context of phylogenetic comparative methods, we hereby focus on the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
 
@@
         Panel~B: Theoretically, changes of a trait under a moving optimum should be better predicted by a chronogram.
-        Instead, changes of a neutral trait should be better predicted a phylogram.
+        Instead, changes of a neutral trait should be better predicted by a phylogram.
     }
@@
 % The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
-Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
-Theoretically, for a neutrally evolving trait, trait changes depend directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the average time between two consecutive generations, called generation time, is constant and the same for each species.
-Since generation time depends on the time for an individual to reach sexual maturity, it vary considerably between species (e.g.
-from a few months for coral reef pygmy gobies to 150 years for Greenland shark).
+Typically, the tree is assumed to be known and obtained independently, and the \textit{de facto} standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
+Theoretically, for a neutrally evolving trait, trait changes depend directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the average time between two consecutive generations, called generation time, is constant and equal for all species.
+Since generation time depends on the time for an individual to reach sexual maturity, it can vary considerably between species, from a few months for coral reef pygmy gobies to 150 years for the Greenland shark~\citep{demagalhaes_database_2009, nielsen_eye_2016}.
 As a result, variations of generation time between species means that modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
@@
 Alternatively, on a phylogram, the branch lengths represent another quantity than time, which can be used as the backbone to model trait evolution and can potentially absorb changes in generation time.
-For example, it is known that using phylograms in units of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}, but in this case the phylogram is not expected to absorb changes in generation time.
+For example, it is known that using phylograms in units of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}.
 From a genomic perspective, phylograms can also be in units of nucleotide divergence, that is, depicting the number of nucleotide substitutions occurring along a branch.
@@
 Additionally, the rational to use nucleotide divergence to absorb changes in generation time is also valid for changes in mutation rate (per loci and per generation) under the assumption of a constant genetic architecture of the trait and that changes in mutation rate impact the whole genome~\citep{latrille_detecting_2024}.
-Empirically, such changes in mutation rate along a phylogenetic tree is also observed, although to a lesser extent than changes in generation time~\citep{bergeron_evolution_2023}.
+Empirically, such changes in mutation rate along a phylogenetic tree are also observed, although to a lesser extent than changes in generation time~\citep{bergeron_evolution_2023}.
 
 % Phylogram in the case of selection
-Under the alternative scenario of selection, shorter generation time and higher mutation rate can also allow for the species to track the changing optimum faster and not lag behind~\citep{lande_natural_1976, hansen_comparative_2008}.
-However, the timescale for this lag (hundred or thousands of years) is negligible compared to timescale on which the optimum changes at the level of clades (millions of years) in first approximation~\citep{hansen_three_2024}.
+Under the alternative scenario involving selection, shorter generation time and higher mutation rate can also allow for the species to track the changing optimum faster and not lag behind~\citep{lande_natural_1976, hansen_comparative_2008}.
+However, the timescale for this lag (hundreds or thousands of years) is negligible compared to the timescale on which the optimum changes at the level of clades (millions of years) in a first approximation~\citep{hansen_three_2024}.
 Thus, at the phylogenetic scale, since shifts in the optimum fitness peak is extrinsic to the species and is dependent on the environment which varies with time, using a chronogram would be more accurate to model mean trait changes.
-As a result, for a trait under selection, mean trait changes should be better explained by the time rather than the number of substitutions that occurred along a branch.
+As a result, for a trait under selection, mean trait changes should be better explained by the time rather than the number of substitutions that occurred along a branch (Fig.~\ref{fig:methods}B).
 
 % Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
-In this study, we test if using a phylogram (in units of nucleotide divergence for neutral loci) instead of a chronogram can allow to assess the regime of evolution of a trait (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
+In this study, we test this hypothesis, namely if using a phylogram (in units of nucleotide divergence for neutral loci) instead of a chronogram (in units proportional to time) can allow to assess the regime of evolution of a trait, by evaluating the fit of a BM on both trees.
 More precisely, we seek to test if the changes of a neutral trait are better predicted by a phylogram.
@@
 \section*{Methods}
-In order to assess whether a phylogram is better suited to model neutral evolution we first performed simulations of a trait evolving along a phylogenetic tree under different regimes of selection (neutral, moving optimum, multiple optima).
-Second, we gathered an empirical dataset of body and brain masses from mammals, including both a chronogram and a phylogram (in unit of nucleotide divergence from neutral loci).
-Third, we fitted a single rate Brownian motion (BM) to the simulated data, and we compared the fit of the BM on a phylogram versus a chronogram.
-Fourth, based in the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated and empirical datasets.
-Finally, we fitted several models of trait evolution to the simulated and empirical datasets: a multi-rate BM, an Ornstein-Uhlenbeck (OU) process, and a relaxed OU with multiple optima.
+To assess whether a phylogram is better suited to model neutral evolution we first performed simulations of a trait evolving along a phylogenetic tree under different regimes of selection (neutral, moving optimum, multiple optima).
+Second, we fitted a single rate Brownian motion (BM) to the simulated data, and we compared the fit of the BM on a phylogram versus a chronogram.
+Third, based on the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated dataset.
+Fourth, we fitted several models of trait evolution to the simulated datasets: a multi-rate BM, an Ornstein-Uhlenbeck (OU) process, and a relaxed OU with multiple optima, tested on a phylogram versus a chronogram.
+Finally, we gathered an empirical dataset of body and brain masses from mammals, including both a chronogram and a phylogram (in units of nucleotide divergence from neutral loci), on which we tested the support of a phylogram for the single rate BM.
 
@@
 The parameters of simulations were chosen to mimic an empirical dataset of mammals.
-In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
+In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation.
 Parents were selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
-Flattening the fitness landscape resulted in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
-Alternatively, for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
+Flattening the fitness landscape resulted in neutral evolution, which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
+Alternatively, for a trait under selection around a moving optimum, we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
 Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch; if a switch occurs, the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).
 
-At each node of the tree, the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached (Fig.~\ref{fig:results-brownian}C).
+At each internal node of the tree (i.e.
+speciation event), the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached.
 As a control, we performed simulations with constant mutation rates, generation times and effective population sizes ($\Ne$).
@@
 
-The phylogram is obtained from the same simulation setting on a set of independent 30000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
-The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
-
-\subsection*{Empirical dataset}
-We analysed a dataset of body and brain masses from mammals.
-The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
-We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
-We also extracted brain and body masses from the \textit{COMBINE} dataset~\citep{soria_combine_2021}.
-The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
-More specifically, the phylogram in units of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
+The phylogram is obtained from the same simulation setting on a set of 30,000 independent neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
 The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
@@
 We used a log-uniform prior for $\sigma$.
-Using Markov chain Monte Carlo (MCMC), we reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
+Using Markov chain Monte Carlo (MCMC), we reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10,000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
 Both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
-Importantly, the data and the tree topology are the same in both analyses, only the branch lengths are different.
-For traits simulated under neutral evolution, we expect more accurate trait reconstruction on a phylogram, and conversely, for traits under selection (moving optimum), more accurate trait reconstruction on a chronogram.
+Importantly, the data and the tree topology are the same in both analyses; only the branch lengths are different.
+In this simulation setting, for traits simulated under neutral evolution, we expect more accurate trait reconstruction on a phylogram, and conversely, for traits under selection (moving optimum), more accurate trait reconstruction on a chronogram.
 
 \subsection*{BM with a switch}
-To test the support of a phylogram over a chronogram for the simulated and empirical datasets, we implemented a model based on a single rate BM.
-The model was implemented in \textit{RevBayes} and contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
-Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
-If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
+To test the support of a phylogram over a chronogram, we implemented a model based on a single rate BM.
+The model was implemented in \textit{RevBayes} and contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions (Fig.~\ref{fig:results-brownian}A).
+Formally, $\PhyloSupport \in {0, 1}$ is a Bernoulli random variable ($\PhyloSupport \in {0, 1}$) with a prior probability of $0.5$.
+If $\PhyloSupport$ is $0$, the branch lengths are those of the chronogram, and if $\PhyloSupport$ is $1$, the branch lengths are those of the phylogram.
 The tree topology is fixed, meaning both trees have the same branching structure, but because branch lengths are different and not necessarily on the same scale, both trees are re-scaled by dividing branch lengths by the total tree length.
@@
 Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
-The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability that the phylogram is favoured over the chronogram.
+The posterior mean of $\PhyloSupport$ (burn-in of 1,000 gen., running of 10,000 gen., 2 chains) is the probability that the phylogram is favoured over the chronogram.
 We expect $\PhyloSupport = 1$ for a trait under neutral evolution, and $\PhyloSupport = 0$ for a trait under selection (moving optimum).
@@
 \subsection*{Alternative models of evolution}
-In addition to a single rate BM, we  also fitted a BM with multiple rate parameters~\citep[auteur]{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
+In addition to a single rate BM, we also fitted a BM with multiple rate parameters~\citep{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
 The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
 For each branch, we draw a rate-multiplier either equal to $1$ (no rate shift), or drawn from a log-normal distribution with a median of 1, and a standard deviation such that rate shifts range over about one order of magnitude.
-We set to our prior expected number of rate shifts to $1$ across the whole tree.
-The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1000 gen., running of 50000 gen., 2 chains).
+We set our prior to the expected number of rate shifts of $1$ across the whole tree.
+The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1,000 gen., running of 50,000 gen., 2 chains).
 
@@
 Thus, when $\alpha = 0$ the OU process (eq.~\ref{eq:ou-process}) is equivalent to a BM (eq.~\ref{eq:bm-process}).
-The support for the OU model is estimated as the posterior mean of the switch variable ($p_{\text{OU}}$), or equivalently that of $\alpha \neq 0$ (burn-in of 1000 gen., running of 10000 gen., 2 chains).
+The support for the OU model is estimated as the posterior mean of the switch variable ($p_{\text{OU}}$), or equivalently that of $\alpha \neq 0$ (burn-in of 1,000 gen., running of 10,000 gen., 2 chains).
 
@@
 Finally, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
-The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1000 gen., running of 5000 gen., 2 chains).
+The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1,000 gen., running of 5,000 gen., 2 chains).
 For each branch, we draw a shift in optimum either equal to $0$ (no shift), or drawn from a uniform distribution centred on $0$ with breadth spanning the whole range of observed trait value.
-We set to our prior expected number of optimum shifts to $1$ across the whole tree.
+We set our prior to the expected number of optimum shifts of $1$ across the whole tree.
+
+\subsection*{Empirical dataset}
+We analysed a dataset of body and brain masses from mammals.
+The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
+We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
+We also extracted brain and body masses from the \textit{COMBINE} dataset~\citep{soria_combine_2021}.
+The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
+More specifically, the phylogram in units of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
+The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.
 
@@
         Panel~A:
-        Test for the support of a phylogram over a chronogram using a Brownian motion (BM) to model trait changes along a phylogeny.
-        The model contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time ($\PhyloSupport=0$, chronogram in purple) to units of nucleotide divergence for neutral sites ($\PhyloSupport=1$, phylogram in green).
-        The posterior mean estimate of $\PhyloSupport$ indicates whether a BM supports better a phylogram or a chronogram ($0 \leq \PhyloSupport \leq 1$), given the mean trait value for each extant species and both trees (chronogram \& phylogram) with the same topology.
-        Only if the phylogram has more support than chronogram ($\PhyloSupport$ close to $1$) there is some support that the trait might evolve under drift.
+        Brownian motion (BM) is used to model trait changes along a phylogeny, given a mean trait value for extant species.
+        The model contains a switch variable, denoted $\PhyloSupport$, that switches the branch lengths in units of time ($\PhyloSupport=0$, chronogram) to units of nucleotide divergence for neutral sites ($\PhyloSupport=1$, phylogram).
+        The posterior mean estimate of $\PhyloSupport$ indicates whether a BM supports better the phylogram over the chronogram ($0 \leq \PhyloSupport \leq 1$).
         Panel~B: Wright-Fisher simulations of trait evolution along a phylogeny.
+        At each node of the tree a speciation event creates two descendant species evolving independently.
         Neutral evolution is modelled as a constant fitness regardless of the phenotype of individuals (yellow).
         Selection is modelled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree: as a moving optimum (blue) or as multiple optima (red).
-        At each node of the tree we have a speciation event and the two descendant species evolve independently.
-        The output of the simulation is the mean trait value for each extant species, but we also save the phylogram and the chronogram as obtained from independent markers.
-        We then compute the posterior mean estimate of $\PhyloSupport$ from the output of the simulation.
+        The output of the simulation is: 1) the mean trait value for each extant species, 2) the phylogram as obtained from independent neutral markers and 3) the chronogram derived from the phylogram assuming a relaxed molecular clock.
+        The posterior mean estimate of $\PhyloSupport$ is inferred from the output of the simulation.
         Panels C \& D: Violin plot of the posterior mean of $\PhyloSupport$ across 100 replicates for the different simulated regimes of evolution with mean $\PhyloSupport$ values at the center ($\MeanPhyloSupport$).
@@
 % Simulations of a trait evolving neutrally versus under a moving optimum.
-We first tested the hypothesis that a phylogram is better suited to model neutral evolution than a chronogram.
-We simulated traits evolving on a phylogenetic tree under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum, and 3) under multiple optima (see Methods, Fig.~\ref{fig:results-brownian}A-C).
-We first assessed the accuracy of ancestral trait reconstruction on a phylogram versus a chronogram, under the assumption that changes follow a Brownian motion (BM).
-For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
+We tested the hypothesis that a phylogram is better suited to model neutral evolution than a chronogram.
+We simulated traits evolving on a phylogenetic tree under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum, and 3) under multiple optima (see Methods).
+We first assessed the accuracy of ancestral trait reconstruction on a phylogram versus a chronogram, under the assumption that changes follow a BM.
+For a neutral trait, ancestral trait reconstruction was more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
 In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$) than on a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).
 
-We then tested for the BM support of phylogram over a chronogram, and we developed as statistic denoted as $\PhyloSupport$, which ranges from 0 (full support for the chronogram) to 1 (full support for the phylogram).
-For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch lengths are equivalent in both trees.
+We then tested for the BM support of phylogram over a chronogram, and we developed a statistic denoted as $\PhyloSupport$ (Fig.~\ref{fig:results-brownian}A), which ranges from 0 (full support for the chronogram) to 1 (full support for the phylogram).
+For simulations with a constant generation time, the support is similar for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}B-C), which is expected since the branch lengths are equivalent in both trees.
 Next, we simulated changing generation time, mutation rate (per generation) and effective population size ($\Ne$) along the phylogenetic tree, mimicking a mammalian range of changes.
-In this case, for a simulated neutral trait, the BM was better fitted on a phylogram than on a chronogram, with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold (Fig.~\ref{fig:results-brownian}E, yellow).
-Conversely, for a simulated trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}E, blue) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}E, red).
+In this case, for a simulated neutral trait, the BM was better fitted on a phylogram than on a chronogram, with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80$ of the $100$ replicates above the $0.95$ threshold (Fig.~\ref{fig:results-brownian}D, yellow).
+Conversely, for a simulated trait under selection, the fit of a BM on a chronogram was better than on a phylogram with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69$ out of $100$ replicates below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}D, blue) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67$ of the $100$ below the $0.05$ threshold, Fig.~\ref{fig:results-brownian}D, red).
 
@@
 % Mainly testing BM versus OU
-Moreover, we tested alternatives to the BM with Ornstein-Uhlenbeck (OU) models: with a single optimum and with multiple optima (see Methods).
+Moreover, we tested alternatives to the BM with OU models: with a single optimum and with multiple optima (see Methods).
 First, for a trait under a moving optimum, the estimated support for the OU process ($p_{\text{OU}}$) was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}C, blue violins, Wilcoxon paired test with $p_{\text{value}}=3.9\times 10^{-18}$), an expected result since the BM with a switch supported the chronogram.
@@
         Posterior estimate for the variance in rate parameters (Panel~A) and number of rate changes (Panel~B).
-        Panel~C: Relative fit an Ornstein-Uhlenbeck (OU) process compared to BM, using either a phylogram or a chronogram.
-        Posteriors estimates for the support of the OU process over the BM (reversible-jump).
+        Panel~C: Relative fit of an Ornstein-Uhlenbeck (OU) process compared to BM, using either a phylogram or a chronogram.
+        Posterior estimates for the support of the OU process over the BM (reversible-jump).
         Panel~D: Fit of a relaxed OU with multiple optima, using either a phylogram or a chronogram.
@@
 However, we argue that to model a trait evolving neutrally, the backbone tree should not be a chronogram but instead a phylogram.
-Phylograms would better represent the number of generations that occurred, and in turn should explain best the differences in traits between species~\citep{latrille_detecting_2024}
+Phylograms better represent the number of generations that occurred, and in turn should explain best the differences in traits between species~\citep{latrille_detecting_2024}
 In practice, we show that for simulations of a trait evolving neutrally, the fit of a BM supports a phylogram rather than a chronogram.
@@
 We applied this method to different empirical datasets of body and brain masses in mammals, and we showed that the chronogram was favoured over the phylogram, ruling out the neutral model of evolution for these traits~\citep{latrille_detecting_2024}.
-The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree (e.g. from 6 months for mice to 50 years for bowhead whales).
+The mammalian dataset is particularly relevant, since changes in generation time are expected to occur along the phylogenetic tree, from 6 months for mice to 50 years for bowhead whales~\citep{demagalhaes_database_2009, nielsen_eye_2016}.
 But more generally, our method is also relevant when changes in mutation rate (per generation) and effective population size ($\Ne$) are expected.
@@
 This argument was partially tested in our simulation by modelling changes in both $\Ne$ and generation time (although not modelling population structure \textit{per se}), and the phylogram was favoured over the chronogram for a trait evolving neutrally.
-Second, under certain conditions, the use of phylograms can also absorbs changes in mutation rate (per loci per generation), which in mammals can vary up to 10 fold~\citep{bergeron_evolution_2023}.
+Second, under certain conditions, the use of phylograms can also absorb changes in mutation rate (per loci per generation), which in mammals can vary up to 10 fold~\citep{bergeron_evolution_2023}.
 As in the case of our simulations, we assumed that the genetic architecture of the trait is constant and the mutation rate is homogeneous across the genome.
 However, if the mutation rate would only change for a subset of the genome that is encoding the trait, the phylogram would not absorb these changes in mutation rate.
-Finally, because chronogram are usually derived from phylogram by calibrating the tree with molecular clocks, using phylograms directly has the additional advantage of removing model assumptions required to calibrate node ages~\citep{litsios_effects_2012}.
+Finally, because chronograms are usually derived from phylograms by calibrating the tree with molecular clocks, using phylograms directly has the additional advantage of removing model assumptions required to calibrate node ages~\citep{litsios_effects_2012}.
 
@@
 These examples show that while phylograms are useful to model neutral evolution, they are not a silver bullet to disentangle the effect of drift and selection.
-In our simulations, we found that generally using a phylogram tends to favour the OU process over the BM model.
+In our simulations, we generally found that using a phylogram tends to favour the OU process over BM.
 This suggests that trees in units of nucleotide divergence may not always provide a reliable framework for modelling trait evolution, especially when selection is involved.
@@
 Caution is indeed warranted, as there are several limitations to our approach.
-First, we simulated only a subset of different regimes of selection, but more complex scenarios could be included, for example with primary and secondary moving optimums~\citep{hansen_three_2024}.
+First, we simulated only a subset of different regimes of selection, but more complex scenarios could be included, for example with primary and secondary moving optima~\citep{hansen_three_2024}.
 Additionally, the simulated genetic architecture of a single trait is assumed to be constant across the phylogeny.
 Constant architecture results in the saturation of the phenotype on long time scales, with a phenomenon similar to nucleotide saturation~\citep{latrille_detecting_2024}.
-This saturation effect might be one of the reasons for a bias in favour of the Ornstein-Uhlenbeck (OU) process over the BM model that we observed in our simulations.
+This saturation effect might be one of the reasons for a bias in favour of the OU process over BM that we observed in our simulations.
 A more realistic genetic architecture should include pleiotropy, epistasis and linkage disequilibrium, while here we only considered non-linked loci contributing additively to a single trait.
 Second, while the tested alternative models of evolution (multi-rates BM, OU, multi-optimum OU) are standard, they do not encompass the full breadth of available methods to model trait evolution~\citep{pennell_geiger_2014, hohna_revbayes_2016}.
-Finally, the mammalian dataset (brain and body) may not be representative of all traits or species and even though our analysis showcase the utility of phylograms, more empirical studies are needed to replicate the findings.
+Finally, the mammalian dataset (brain and body) may not be representative of all traits or species and even though our analysis showcases the utility of phylograms, more empirical studies are needed to replicate the findings.
 
 % Gene expression level and empirical studies.
-One striking example of a trait that could (and should) be empirically studied with a phylogram is gene expression level, both for protein concentration and mRNA expression.
+One striking example of a trait that could (and should) be empirically studied with a phylogram is gene expression level, both for mRNA and protein levels.
 Indeed, the regime of selection acting on gene expression level is the focus of intense debate~\citep{signor_evolution_2018, price_detecting_2022, bertram_cagee_2023, dimayacyac_evaluating_2023}.
-Typically, datasets on which the BM is favoured over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
+Typically, datasets on which BM is favoured over other models is interpreted as mRNA expression level evolving neutrally~\citep{khaitovich_evolution_2006, catalan_drift_2019, dimayacyac_evaluating_2023}.
 Phenotypic distance is also plotted as a function of time while claiming that a trait is evolving under drift~\citep{jiang_decoupling_2023}.
-Instead, we argue that phenotypic distance should be shown as the function of branch length in unit of substitutions (square root), not time.
-More generally, from an empirical perspective, the use of substitutions at neutral loci brings new perspective, particularly when variation within species is also available and the neutrality index for a trait can be computed~\citep{latrille_detecting_2024}.
+Instead, we argue that phenotypic distance should be shown as a function of branch length in units of substitutions (square root), not time.
+More generally, from an empirical perspective, the use of nucleotide changes at neutral loci as a normalizing factor brings new perspectives into trait evolution~\citep{latrille_detecting_2024}.
 
@@
 We provide a Bayesian model implemented in \textit{RevBayes} to test this hypothesis, which indicates whether a BM supports better a phylogram or a chronogram ($0 \leq \PhyloSupport \leq 1$), given the data at the tip of the tree and both trees with the same topology.
-Only if the phylogram has more support than chronogram ($\PhyloSupport$ close to $1$) there is some support that the trait might evolve under drift, but caution is still warranted.
+Only if the phylogram has more support than the chronogram ($\PhyloSupport$ close to $1$) can it be assumed that the trait is evolving under drift, but caution is still warranted.
 
@@
 \section*{Funding}
-Université de Lausanne
+This work was funded by Faculté de Biologie et de Médecine, Université de Lausanne (https://www.unil.ch; to TL, TG and NS) and Swiss National Science Fund (https://www.snf.ch; grant 315230-219757 to TL and NS).
+The funders did not play any role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 
@@
 \subsection{Simulation along a tree}\label{subsec:simulation-along-a-tree}
-% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
+% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5,000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
 % pS = 0.0027577478571428568
@@
 At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
-Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
+Selection was modelled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
 More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
@@
 \newpage
+\section{Ancestral trait reconstruction}\label{sec:ancestral-trait-reconstruction}
+\begin{center}
+    \begin{minipage}{0.49\linewidth}
+        \flushleft {\tiny  \textbf{A}}
+        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_NeutralPhylo}
+    \end{minipage}
+    \begin{minipage}{0.49\linewidth}
+        \flushleft {\tiny  \textbf{B}}
+        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_NeutralChrono}
+    \end{minipage}\\
+    \begin{minipage}{0.49\linewidth}
+        \flushleft {\tiny  \textbf{C}}
+        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_SelectionPhylo}
+    \end{minipage}
+    \begin{minipage}{0.49\linewidth}
+        \flushleft {\tiny  \textbf{D}}
+        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_SelectionChrono}
+    \end{minipage}
+    \captionof{figure}{
+    \textbf{Phenotypic distance as a function of branch length.}
+    Phenotypic distance ($|\Delta z|$) is computed as the absolute difference of the posterior mean trait value between the ancestral node and the descendent node.
+    Across all simulated dataset (100 replicates), we computed the mean phenotypic distance (blue dots) and standard deviation (blue horizontal lines) for each branch.
+    Simulations under a neutral regime (yellow, top row, panels A and B) and under a moving optimum regime (bottom row, panels C and D).
+    Reconstruction of ancestral trait on a phylogram (blue, left column, panels A and C) and a chronogram (right column, panels B and D).
+    \label{fig:results-distance}
+    }
+\end{center}
 
+\newpage
 \section{Empirical dataset}\label{sec:empirical-dataset}
@@
 
-\newpage
-\section{Ancestral trait reconstruction}\label{sec:ancestral-trait-reconstruction}
-\begin{center}
-    \begin{minipage}{0.49\linewidth}
-        \flushleft {\tiny  \textbf{A}}
-        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_NeutralPhylo}
-    \end{minipage}
-    \begin{minipage}{0.49\linewidth}
-        \flushleft {\tiny  \textbf{B}}
-        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_NeutralChrono}
-    \end{minipage}\\
-    \begin{minipage}{0.49\linewidth}
-        \flushleft {\tiny  \textbf{C}}
-        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_SelectionPhylo}
-    \end{minipage}
-    \begin{minipage}{0.49\linewidth}
-        \flushleft {\tiny  \textbf{D}}
-        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_dist_SelectionChrono}
-    \end{minipage}
-    \captionof{figure}{
-    \textbf{Phenotypic distance as a function of branch length.}
-    Phenoypic distance ($|\Delta z|$) is computed as the absolute difference of the posterior mean trait value between the ancestral node and the descendent node.
-    Across all simulated dataset (100 replicates), we computed the mean phenotypic distance (blue dots) and standard deviation (blue horizontal lines) for each branch.
-    Simulations under a neutral regime (yellow, top row, panels A and B) and under a moving optimum regime (bottom row, panels C and D).
-    Reconstruction of ancestral trait on a phylogram (blue, left column, panels A and C) and a chronogram (right column, panels B and D).
-    \label{fig:results-distance}
-    }
-\end{center}
-
 %TC:endignore

### 2025-08-14 manuscript/compile.sh

@@ -1,0 +1,1 @@
+latexdiff main-first-submission.tex main.tex > diff.tex

### 2025-11-17 manuscript/main.tex
Formalism rational draft

@@
 \newcommand{\Multiply}{\cdot}
+\DeclareMathOperator{\E}{\mathbb{E}}
 \DeclareMathOperator{\Var}{\text{var}}
+\DeclareMathOperator{\Cov}{\text{cov}}
 \newcommand{\der}{\mathrm{d}}
@@
 \newcommand{\brownian}{\mathcal{B}}
+\newcommand{\proba}{\mathbb{P}}
+\newcommand{\pfix}{\proba_{\text{fix}}}
+\newcommand{\Spi}{i}
+\newcommand{\Spj}{j}
+\newcommand{\NbrTaxa}{n}
+\newcommand{\Time}{T}
+\newcommand{\NbrGen}{g_{\Spi, \Spj}}
+\newcommand{\NucDiv}{d_{\Spi, \Spj}}
+\newcommand{\VarGenetic}{V_{\mathrm{A}}}
+\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
+\newcommand{\MeanTrait}{\bar{\Trait}}
+\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
+\newcommand{\RootTrait}{\phi}
+\newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
+\newcommand{\VecZero}{\UniDimArray{0}}
+\newcommand{\VecOne}{\UniDimArray{1}}
+\newcommand{\Distance}{\BiDimArray{D}}
+\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
+\newcommand{\SubRate}{q}
+\newcommand{\VarPhenotype}{V_{\Trait}}
+\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
+\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
+\newcommand{\VarMutation}{V_{\mathrm{M}}}
+\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
+\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
+\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
+\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
+\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
+\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
+\newcommand{\EstRateBetween}{\widehat{\sigma}^2_{\mathrm{B}}}
+\newcommand{\EstRateWhithin}{\widehat{\sigma}^2_{\mathrm{W}}}
+\newcommand{\NI}{\rho}
+\newcommand{\EstNI}{\widehat{\rho}}
+\newcommand{\StdSelection}{\sigma}
+\newcommand{\VarSelection}{\StdSelection^2}
 
@@
 
+\section{Theoretical rationale}
+
+For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
+
+For a given species $\Spi$, we denote by $\MeanTrait_{\Spi}$ the mean value of the trait across the individuals.
+If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
+Given a phylogenetic tree, for a pair of species $\Spi$ and $\Spj$ from this tree, we denote as $\NbrGen$ the number of generations between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \citet{hansen_translating_1996} by:
+\begin{align}
+    \VarPhy  = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar}.
+\end{align}
+Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
+The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$~\citep{kimura_probability_1962}.
+We can derive the substitution rate ($\SubRate$, per nucleotide site per generation) as the number of newly arisen mutations ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{kimura_evolutionary_1968}, giving:
+\begin{align}
+    \SubRate & = 2 \Ne \Multiply \MutationRateNuc \Multiply \pfix, \\
+    & = 2 \Ne  \Multiply \MutationRateNuc  \Multiply \dfrac{1}{2\Ne}, \\
+    & = \MutationRateNuc. \label{eq:substitution-rate}
+\end{align}
+That is, if mutations are neutral, the rate of substitution per generation within a genomic region equals the rate at which new mutations arise per generation for the same genomic region, reviewed in \citet{mccandlish_modeling_2014}.
+
+Next, we denote $\NucDiv$ as the nucleotide divergence between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
+In other words, $\NucDiv$ is the expected number of substitutions per nucleotide site during the $\NbrGen$ generations.
+Assuming that no multiple substitutions occurred at the same site, $\NucDiv$ is the number of generations ($\NbrGen$) multiplied by the nucleotide substitution rate per generation ($\SubRate$):
+\begin{align}
+    \NucDiv & = \NbrGen \Multiply \SubRate \\
+    & = \NbrGen \Multiply \MutationRateNuc \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
+\end{align}
+
+We show that the covariance in mean trait value between a pair of species ($\VarPhy$) does increase linearly with shared nucleotide divergence ($\NucDiv$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
+Importantly, since the number of generations is the ratio of time divided by generation time (average time between two consecutive generations), removing the effect of the number of generations also removes the effect of both time and generation time.
+
 \section{Simulation of continuous trait}\label{sec:simulator}
